### 2024-04-07 draft.Rmd
added draft file, picture, and data

---
title: "Draft"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

![Female Willow Ptarmigan captured by [Bryce W. Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](./assets/ptarmigan.jpg)

# Introduction

## Related Work

## Data

# Data Analysis

```{r libraries}
library(readxl)
library(ggplot2)
library(knitr)
```

```{r data_table}
data <- read_excel("./data/bird_data.xlsx")

knitr::kable(data[1:5, 1:5], caption = "Data")
```

```{r}

```

# Methods

# Results

# Conclusion

### 2024-04-12 draft.Rmd
added asset image, made changes to .bib and draft

@@
 ---
-title: "Draft"
+title: "Investigating the alternative prey hypothesis with the POMP framework"
 author: "Pete, Sizhuang, and Heyuan"
 date: "`r Sys.Date()`"
+csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
 output: 
-  html_document:
+  bookdown::html_document2:
     theme: cosmo
@@
 
+Studying ecology, especially animal population dynamics, is essential for understanding the complex interactions between species and their environment, guiding conservation efforts, and ensuring the sustainable use of natural resources for future generations..
+By modeling population dynamics, researchers and scientists can understand how animals respond to environmental changes and human activities.
+This helps conservation efforts by identifying threatened species and develop strategies to protect biodiversity.
+Moreover, animal populations can serve as reservoirs for infectious diseases that can affect both human and animal health.
+Having a robust understanding of population movement can help predict disease outbreaks, understand transmission patterns, and improve disease control and prevention.
+
+This report extends previous research investigating the long-term population decline of willow ptarmigan *Lagopus lagopus*.
+
+- The alternative prey hypothesis predicts that the interaction between generalist predators and their main prey is a major driver of population dynamics of alternative prey species.
+
+- Grouse species are of the most studied
+
+- In this report, we analyzed 142 years of data to investigate whether the population growth of a declining alternative prey species, the willow ptarmigan, is related to an interaction between red foxes and their main prey, small rodents.
+
 ## Related Work
@@
 
+Our data includes 142 observations of CPUE from 1872 to 2012.
+The harvest data was recorded by hunters in the Southeastern mountain regions of Norway.
+The data is provided by the authors of the study and is hosted by Dryad [here]().
+A description of the dataset provided by the authors is given below:
+
+- **year**: Self explanatory
+- **CPUE**: Catch per unit effort, expressed as a number of birds shot per hunter per year.
+- **log.CPUE**: Logarithm transformation of CPUE
+- **peak_rodent_year**: Peak rodent year is scored as “yes”, otherwise “no”. Occurrence of peak
+rodent years were extracted from four sources: 1871-1949 (Wildhagen 1952), 1932-1971
+(Myrberget 1982a), 1971-1979 (Christiansen 1983), 1981-1985 (Frafjord 1988), 1986- 1989
+(Selås et al. 2013), and 1990-2013 (Framstad 2017). Note that 1980 is missing as we were
+not able to find an estimate for this year.
+
 # Data Analysis
 
-```{r libraries}
+```{r libraries, echo=FALSE, message=FALSE}
 library(readxl)
@@
 library(knitr)
+library(forecast)
+library(dplyr)
+library(cowplot)
+library(pomp)
 ```
 
-```{r data_table}
+```{r data_table, echo=FALSE}
 data <- read_excel("./data/bird_data.xlsx")
+data |>
+  rename(peak_rodent_year = "peak rodent year") -> data
 
-knitr::kable(data[1:5, 1:5], caption = "Data")
+knitr::kable(data[1:5, 1:6], caption = "Data")
 ```
+
+```{r, echo=FALSE, message=FALSE}
+# Most basic bubble plot
+p1 <- ggplot(data, aes(x=year, y=CPUE)) +
+  geom_line(color="steelblue") + 
+  geom_point() + 
+  geom_rect(
+    data = subset(data, peak_rodent_year == 'yes'),
+            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
+            fill = "blue", alpha = 0.3
+    ) + 
+  theme_light() +
+  xlab("") + 
+  stat_smooth(
+    color = "#FC4E07",
+    method = "loess"
+  )
+
+p2 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
+  geom_line(color="steelblue") + 
+  geom_point() + 
+  geom_rect(
+    data = subset(data, peak_rodent_year == 'yes'),
+            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
+            fill = "blue", alpha = 0.3
+    ) + 
+  theme_light() + 
+  xlab("Year") + 
+  stat_smooth(
+    color = "#FC4E07",
+    method = "loess"
+  )
+
+plot_grid(
+  p1, p2,
+  labels = "AUTO", 
+  ncol = 1
+)
+```
+
+Looking at both *CPUE* and *log.CPUE* in [figure], it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
+The blue shaded regions highlight years when peak rodent year is "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
+It's also worth noting that this trend is far less apparent beyond the early 1900s.
+
+In accordance with the previous study, our analysis will use *log.CPUE* to reduce heteroscedasticity.
+
+```{r ACF, echo=FALSE}
+acf(data$log.CPUE, lag = 100, main = "ACF log.CPUE")
+```
+
+```{r PACF, echo=FALSE}
+pacf(data$log.CPUE, lag = 100, main = "PACF log.CPUE")
+```
+
+--------
+
+# Methods
+
+## ARMA Baseline
+
+We will start off by fitting a simple ARMA model with parameters chosen using the algorithmic approach, grid search.
+This simple model will serve as a reference against which the performance of the more advanced POMP model is measured.
+Essentially, the ARMA model will serve as a yardstick for managing expectations.
+Of the models shown in (table #), we select the one with the Akaike’s information criterion (AIC).
+AIC is essentially “minus twice the maximized log likelihood plus twice the number of parameters [cite],” and is defined by:
+
+$$AIC=-2\times\ell(\theta^*)+2D$$
+
+```{r grid_search, echo=FALSE, warning=FALSE}
+aic_table <- function(data,P,Q){
+  table <- matrix(NA,(P+1),(Q+1))
+  for(p in 0:P) {
+    for(q in 0:Q) {
+      table[p+1,q+1] <- arima(data,order=c(p,0,q))$aic
+    }
+  }
+  dimnames(table) <- list(paste("AR",0:P, sep=""),
+    paste("MA",0:Q,sep=""))
+  table
+}
+
+bird_aic_table <- aic_table(data$log.CPUE,4,5)
+require(knitr)
+kable(bird_aic_table,digits=2)
+```
+
+## Partially Observed Markov Process
+
+Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models "consist of an unobserved Markov state process, connected to the data via an explicit model of the observation process." The former is referred to as the *latent process model* and the latter as the *measurement model.*
+
+![Conditional independence graph of POMP from A. King's Notes](./assets/state_space.png)
+
+### Ricker Model
 
@@
 
-# Methods
+--------
+
+## Diagnostics
+
+--------
 
@@
 
+--------
+
 # Conclusion
+
+--------
+
+# References

### 2024-04-15 workspace/heyuan.Rmd
Add files via upload

```{r}
library(tidyverse)
library(tseries)
```

#### read in the data & data description
```{r}
data_path <- '/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/data.txt'
df <- read.table(data_path, header = TRUE, sep = "\t", na.strings = "NA", dec = ".", strip.white = TRUE)
head(df)

```

142 years (1872-2013) of willow ptarmigan hunting records in southeastern Norway

CPUE: the number of birds shot per hunter per day, as the primary metric for assessing population changes.

NAO: A positive NAO index favors mild winters in the U.S. East. Negative values favor stronger cold-air outbreaks and increased storminess in the eastern U.S.

temp.PREINC: temperature, pre-incubation, 5/1 - 6/2
temp.INC: temperature, incubation, 6/3 - 6/24
temp.BROOD: temperature, brooding & rearing, 6/25 - 7/15
temp.HATCH: temperature, mean hatching date, 7/21 - 7/27
prefix prec.: precipitation

opening data hunt(julian day): the starting day of hunting season is named as Julian Day, which is varied across 142 years. It is mainly in mid to late August up to the 1920's and 10th/15th September in later years.

# Data Analysis

#### Trend Analysis
Temporal Patterns: CPUE showed significant declines from early records to recent years, with notable shifts in population cycles and their amplitude.

Impact of Alternative Prey: The availability of rodents, which serve as alternative prey for predators that also hunt ptarmigan, was strongly linked to fluctuations in ptarmigan numbers. Years with high rodent populations typically saw higher CPUE for ptarmigan, suggesting lower predatory pressure on them.

#### Correlation
For the ACF plot, the initial 40 lags show a strong correlation with past values, diminishing quickly after first 5 lags. It suggests $q = 5$, $q = 40$ or any value smaller than 40 appropriate for MA(q) component. For the PACF plot, partial correlation dies out quickly after 5 lags. It would be appropriate to select $p = 0$ or $p = 5$ for AR(p) component.

Given the long-range significance in the ACF, I may be dealing with a non-stationary series that needs differencing. Also this plot with a flip to negative autocorrelations and a cyclical pattern that declines over time is a characteristic sign of seasonality in my data. However, STL or classical decomposition fails because time series has no or less than 2 periods, indicating the non-existence of seasonality. So the repeated short-term cycle in 5 lags could be explained by the partial correlation.

```{r, error=TRUE}
start_year <- min(df$year)
end_year <- max(df$year)
frequency <- 1  
ts_cpue <- ts(df$log.CPUE, start=start_year, end=end_year, frequency=frequency)

# Apply STL decomposition
ts_components <- stl(ts_cpue, s.window = "periodic")
plot(ts_components)
```

#### Stationary Check
ARMA(p,q) is performed only if time series data is stationary. So, I will do the stationarity check by ADF.

```{r}
# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss_result <- kpss.test(ts_cpue, null = "Level")
print(kpss_result)
```

From ADF test, p-value is estimated as 0.01, smaller than 0.05. We have enough evidence to reject the null hypothesis that the time series is non-stationary.

KPSS with p-value = 0.01 suggests that the log CPUE times series data is non-stationary. This means I can't fit the data with ARMA(p,q) without differencing to achieve stationarity.

#### Differencing
```{r}
# Differencing the time series once
ts_diff <- diff(ts_cpue, differences = 1)
print(kpss.test(ts_diff,null = "Level"))
```

KPSS with p-value = 0.1 suggests that the first-order differenced times series data is stationary. This means I can fit this differenced data with ARMA(p,q).

# Methods

## ARMA
From ADF test, p-value is estimated as 0.01, smaller than 0.05. We have enough evidence to reject the null hypothesis that the time series is non-stationary. KPSS also support the same conclusion that this times series data is stationary. Thus, we could conclude that CPUE on log scale is stationary. This means I can fit the data with ARMA(p,q) without the need to difference or other ways of data transformation to achieve stationarity.

```{r}

aic_table <- function(data,P,Q){
  table <- matrix(NA,(P+1),(Q+1))
  for(p in 0:P) {
    for(q in 0:Q) {
      table[p+1,q+1] <- arima(data,order=c(p,0,q))$aic
    }
  }
  dimnames(table) <- list(paste("AR",0:P, sep=""),
    paste("MA",0:Q,sep=""))
  table
}

bird_aic_table <- aic_table(ts_diff,5,10)
require(knitr)
kable(bird_aic_table,digits=2)

```

The smaller AIC, the better the model to fit the data. Thus, ARMA(0,5) with the lowest AIC value of 204.48 is the best model.

```{r}
aic_table <- function(data,P,Q){
  table <- matrix(NA,(P+1),(Q+1))
  for(p in 0:P) {
    for(q in 0:Q) {
      table[p+1,q+1] <- arima(data,order=c(p,1,q))$aic
    }
  }
  dimnames(table) <- list(paste("AR",0:P, sep=""),
    paste("MA",0:Q,sep=""))
  table
}

bird_aic_table <- aic_table(ts_cpue,5,10)
require(knitr)
kable(bird_aic_table,digits=2)
```

ARIMA(0,1,5) achieves the same result for original log CPUE data.

## POMP

### 2024-04-16 workspace/heyuan.Rmd
Add files via upload

@@
 library(tseries)
+#install.packages("ggfortify")
+library(ggfortify)
 ```
@@
 
-## POMP
+ARMA(0,5) Diagnosis
+```{r}
+fit1 <- arima(ts_cpue, order=c(0,0,5))
+residuals <- residuals(fit1)
+
+# ACF check
+acf(residuals, main="ACF of Residuals")
+
+# Normality check with Q-Q plot
+qqnorm(residuals)
+qqline(residuals, col="red")
+
+# Shapiro-Wilk test
+shapiro.test(residuals)
+
+# Root Check
+ma_coefficients <- coef(fit1)[grepl("ma", names(coef(fit1)))]
+MA_roots <- polyroot(c(1, ma_coefficients))
+#invertible <- all(Mod(MA_roots) > 1)
+#print(paste("Is the model invertible? ", invertible))
+
+autoplot(fit1) +
+  ggtitle("ARMA(0, 5)")
+```
+
+For the ACF plot of the residuals, all autocorrelations are within the confidence bounds (indicated by blue dashed lines), which suggests that there is no significant autocorrelation present.
+Additionally, the QQ plot shows that the data points roughly align with the reference line (colored red), supporting the assumption of normality.
+This is further confirmed by a Shapiro-Wilk test for normality, which returns a p-value of 0.62, indicating that the residuals can be considered normally distributed.
+Moreover, the roots of the characteristic polynomial are outside the unit circle, confirming that the MA(5) process is invertible—a critical condition for model adequacy.
+These diagnostics collectively verify that the ARIMA(0,1,5) model is well-specified and adheres to the necessary assumptions for valid time series analysis.
+
+## POMP 1
+
+```{r}
+install.packages("pomp")
+library(pomp)
+```
+
+To create a POMP for CPUE, I need to define the state process that reflects the underlying dynamics of the species population and the observation process that relates the state to the CPUE measurements.

### 2024-04-18 workspace/sizhuang.Rmd
sizhuang EDA

--- 2024-04-12 draft.Rmd
@@
 ```{r data_table, echo=FALSE}
-data <- read_excel("./data/bird_data.xlsx")
+data <- read_excel("../data/bird_data.xlsx")
 data |>
@@
 
-knitr::kable(data[1:5, 1:6], caption = "Data")
+data[data == "NA"] <- NA
+clean_data <- na.omit(data)
+
+# knitr::kable(clean_data[1:5, 1:6], caption = "Cleaned Data")
+```
+
+```{r}
+columns_to_convert <- c("growth rate CPUE", "temp.PREINC", "temp.INC", "temp.HATCH", "temp.BROOD", "temp.WINTER", "temp.SPRING", "prec.PREINC", "prec.INC", "prec.HATCH", "prec.BROOD")
+# Convert these columns to numeric
+clean_data[columns_to_convert] <- lapply(clean_data[columns_to_convert], as.numeric)
+
+# Check for any remaining NAs
+# sapply(clean_data[columns_to_convert], function(x) sum(is.na(x)))
+```
+
+```{r}
+# Summary statistics for all numerical variables
+summary(clean_data)
+
 ```
@@
 # Most basic bubble plot
-p1 <- ggplot(data, aes(x=year, y=CPUE)) +
+p1 <- ggplot(clean_data, aes(x=year, y=CPUE)) +
   geom_line(color="steelblue") + 
@@
   geom_rect(
-    data = subset(data, peak_rodent_year == 'yes'),
+    data = subset(clean_data, peak_rodent_year == 'yes'),
             aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
@@
 
-p2 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
+p2 <- ggplot(clean_data, aes(x=year, y=log.CPUE)) + 
   geom_line(color="steelblue") + 
@@
 ```{r ACF, echo=FALSE}
-acf(data$log.CPUE, lag = 100, main = "ACF log.CPUE")
+acf(clean_data$log.CPUE, lag = 100, main = "ACF log.CPUE")
 ```
@@
 ```{r PACF, echo=FALSE}
-pacf(data$log.CPUE, lag = 100, main = "PACF log.CPUE")
+pacf(clean_data$log.CPUE, lag = 100, main = "PACF log.CPUE")
+```
+
+```{r}
+# Box plot of CPUE for each Peak Rodent Year
+ggplot(data, aes(x = peak_rodent_year, y = CPUE, fill = peak_rodent_year)) +
+  geom_boxplot() +
+  labs(title = "CPUE by Peak Rodent Year", x = "Peak Rodent Year", y = "CPUE")
 ```

### 2024-04-19 draft.Rmd
latest changes

@@
   bookdown::html_document2:
+    code_folding: hide
     theme: cosmo
@@
 
-- In this report, we analyzed 142 years of data to investigate whether the population growth of a declining alternative prey species, the willow ptarmigan, is related to an interaction between red foxes and their main prey, small rodents.
+- In this report, we analyze 142 years of data to investigate whether the population growth of a declining alternative prey species, the willow ptarmigan, is related to an interaction between red foxes and their main prey, small rodents.
+
+Two types of questions we can ask.
+
+- Scientific goals: Which explanations are most favored by the data? Which kinds of data are most informative?
+- Applied goals: how to design ecological or epidemiological intervention? How to make accurate forecasts?
+
+This report seeks to answer two scientific questions.
+First, between alternative prey hypothesis and the role of climate change, which explanation is most favored by the data?
+Secondly, which kind of data is more informative?
 
 ## Related Work
+
+In ecological studies, researching predator-prey dynamics, especially within the context of the alternative prey hypothesis, offers profound understanding into ecosystem complexities.
+Focusing on ptarmigans in Norway, a number of studies have attempted to model the interplay between ptarmigans as prey and their predators, including gyrfalcons and foxes.
+Notably, research by Aanes et al.
+(2002) highlighted the significant impact of predation pressure on ptarmigan populations, emphasizing the need to understand predator-prey dynamics for effective conservation strategies.
+
+## Objectives
+
+Our research objectives are as follow:
+
+- 
+- There has been a lack in application of the Partially Observed Markov Process (POMP) framework to ecological research questions.
+
+The remainder of the report is structured as follows.
+We briefly introduce and describe our data set and variables of interest.
+Following that, a short exploration of data trend, correlation, and association is is given.
+Next, we build a baseline ARMA process model to serve as a measurement for our more complex proposals.
+We then introduce the pomp framework and present our model along with a section on model diagnostics.
+We conclude with the results, a discussion, and a summary with suggestions of future extensions.
 
@@
 not able to find an estimate for this year.
-
-# Data Analysis
+- **NAO**: The mean North Atlantic Oscillation index for the months May through
+July.
 
@@
 data |>
-  rename(peak_rodent_year = "peak rodent year") -> data
+  rename(peak_rodent_year = "peak rodent year") |>
+  rename(NAO = "NAO May_Jun_Jul")-> data
 
-knitr::kable(data[1:5, 1:6], caption = "Data")
+knitr::kable(data[1:5, 1:7], caption = "Data")
 ```
+
+# Data Analysis
+
+## Trend Analysis {.tabset}
+
+### Time Plots
 
@@
   geom_point() + 
-  geom_rect(
-    data = subset(data, peak_rodent_year == 'yes'),
-            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
-            fill = "blue", alpha = 0.3
-    ) + 
   theme_light() +
@@
 p2 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
+  geom_line(color="steelblue") + 
+  geom_point() + 
+  theme_light() + 
+  xlab("Year") + 
+  stat_smooth(
+    color = "#FC4E07",
+    method = "loess"
+  )
+
+p3 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
   geom_line(color="steelblue") + 
@@
 plot_grid(
-  p1, p2,
+  p1, p2, p3,
   labels = "AUTO", 
@@
 
+### Box Plot
+
+```{r, boxplot, echo=FALSE}
+ggplot(data, aes(x = peak_rodent_year, y = log.CPUE)) +
+
+  geom_boxplot() + 
+  theme_light() + 
+  xlab("Peak Rodent Year")
+```
+
+## {-}
+
 Looking at both *CPUE* and *log.CPUE* in [figure], it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
-The blue shaded regions highlight years when peak rodent year is "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
+The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
 It's also worth noting that this trend is far less apparent beyond the early 1900s.
+
+Figure [] depicts the ACF plot of *log.CPUE* at 100 lags.
+The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
+More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
+The PACF plot in figure [], the partial correlation dies out after 5 lags.
+For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
+
+Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data.
+Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods.
+The repeated short-term cycle in 5 lags could be explained by the partial correlation.
 
@@
 
+## Correlation {.tabset}
+
+### ACF
 ```{r ACF, echo=FALSE}
@@
 
+### PACF
+
 ```{r PACF, echo=FALSE}
@@
 ```
-
+## {-}
 --------
@@
 
-Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models "consist of an unobserved Markov state process, connected to the data via an explicit model of the observation process." The former is referred to as the *latent process model* and the latter as the *measurement model.*
+```{r, echo=FALSE}
+# This code snippet is to add a new column that mutates peak_rodent_year into 
+# binary values for the POMP model.
+data |>
+  mutate(rodent_binary = ifelse(peak_rodent_year == 'yes', 1, 0)) -> data
+```
+
+Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models "consist of an unobserved Markov state process, connected to the data via an explicit model of the observation process." The former is referred to as the *latent process model* and the latter as the *measurement model.* A simple visual representation of the process is depicted in figure [].
+
+We propose modeling the pomp framework using a variation of the Lotka–Volterra equations [cite].
+The Lotka-Voltera equations, also referred to as the predator-prey equations, are a pair of first order nonlinear differential equations.
+They are frequently used to describe the dynamics of biological systems in which two species interact, one as a predator and the other as prey.
+Our analysis focuses on a slightly different approach that introduces the idea of an *alternative prey.* The alternative prey hypothesis supposes that predators supported by a primary species will shift to consume alternative prey during a decrease in primary prey abundance.
+
+The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator and alternative prey (rodent) count.
 
@@
 
-### Ricker Model
+### Model Components
 
-```{r}
+- `rprocess`: a draw from $f_{X_n|X_{n-1}}(x_n|x_{n-1};\theta)$
+- `dprocess`: evaluation of $f_{X_n|X_{n-1}}(x_n|x_{n-1};\theta)$
+- `rmeasure`: a draw from $f_{Y_n|X_{n}}(y_n|x_{n};\theta)$
+- `dmeasure`: evaluation of $f_{Y_n|X_{n}}(y_n|x_{n};\theta)$
+- `rinit`:a draw from $f_{X_0}(x_0;\theta)$
 
+**Covariates**
+
+- $R(t)=$ peak rodent year, from data
+
+$$Y_t \sim NB(\rho B_t, \sigma),$$
+where parameter $\rho$ is proportional to capturing effort, e.g., the number of hunters.
+
+$$F_{t+dt}=\delta(t)F_tB_t-\theta F_t$$ 
+where
+
+$$
+\delta (t) = \left\{\begin{array}{ll}
+\delta_1 & : R(t)=0\\
+\psi \delta_1 & :  o.w.
+\end{array}
+\right.
+$$
+
+$\delta(t)F_t\beta_t$ is the rate of the poisson step.
+
+$$B_{t+dt} = \alpha B_t - \beta(R_t)B_tF_t,$$
+
+where
+
+$$
+\beta (R_t) = \left\{\begin{array}{ll}
+\beta_1 & : R(t)=0\\
+\gamma \beta_1 & :  o.w.
+\end{array}
+\right.
+$$
+
+We use a binomial approximation with transition probabilities
+
+```{r, eval=FALSE}
+Csnippet("
+  double dNf  = rbinom(F, 1-exp);
+  double dF_p = rbinom(F, 1-exp(rho_t * beta_t * dt));
+  double dF_m = rbinom(F, 1-exp(theta*dt));
+
+  B = 
+  F += dF_p - dF_m;
+
+") -> step
+
+Csnippet("
+  B = 
+  F = 
+") -> rinit
+
+Csnippet("
+
+") -> rproc
+
+Csnippet("
+  count = dnbinom(rho*B, sigma)
+") -> rmeas
+
+Cnsippet("
+
+") -> dmeas
+
+mod <- pomp(rprocess = euler(step, delta.t =),
+            times = "year", t0 = 1872,
+            rinit = rinit,
+            rprocess = eurler(step, delta.t=),
+            rmeasure = rmeas, 
+            dmeasure = , 
+            paramnames = c("rho", "theta", "delta"),
+
+)
 ```
@@
 
+# Contribution
+
+- Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
+- Heyuan: ?
+- Sizhuang:
+
+--------
+
 # References

### 2024-04-22 draft.Rmd
minor change

@@
 
-```{r libraries, echo=FALSE, message=FALSE}
+```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
 library(readxl)
@@
 library(cowplot)
-library(pomp)
+#library(pomp)
 ```
@@
   rename(peak_rodent_year = "peak rodent year") |>
-  rename(NAO = "NAO May_Jun_Jul")-> data
+  rename(NAO = "NAO May_Jun_Jul") |>
+  select(c(year, CPUE, peak_rodent_year, NAO)) -> data
 
-knitr::kable(data[1:5, 1:7], caption = "Data")
+knitr::kable(data, caption = "Data")
 ```

### 2024-04-22 workspace/heyuan.Rmd
Add files via upload

@@
 library(ggfortify)
+library(pomp)
 ```
@@
 
-## POMP 1
+## POMP 3
+
+Y(t): bird count, $y_t = NB(mean = \rho B_t, \sigma)$, $B_t$ is bird population size at time t
+X(t): unobserved state process, red foxes, Bird, Peak Rodent Year, temperature, precipitation, NAO index
+
+Reference: Lotka-Volterra predator-prey model for the state process: $$X_{t+dt} = \beta X_t - \mu X_t$$
+
+natural born - number of birds eaten by foxes: 
+$$B_{t+dt} = \alpha B_t - \beta(R_t) B_{t}F_{t}$$
+
+where, $$\beta(R_t) = \beta if R(t)=0$$ or 
+$$ \beta(R_t) = \gamma \beta, \gamma > 1$$ , rodent number affect the rate of fox eating birds, 
+$\beta(R_t)$ can be re-written as $$\beta(R_t) =  \gamma^R \beta$$, where R = 0 or 1
+
+Growth Rate - Death Rate of Foxes: 
+$$ F_{t+dt} = \delta(t) F_t B^*_t - \theta(t) F_t$$
+
+where, $$\delta(t) = \delta_1 if R(t) = 0, or, \phi \delta_1$$ , $\phi<1$ and whether it is in rodent peak year affect the number of birds and the number of foxes.
+
+$\delta(t)$ can be re-written as $$\delta(t) = \phi^R \delta_1$$, where R = 0 or 1
+
+$\theta(t)$ could include the temperature, NAO about how does it affect the death rate of Foxes.
+
+first, let dt=1 year
+then, dt = less than 1 year, maybe 1 month, bird keeps as 100, foxes=10, 15, 17, 15 for each dt until reaching 1 year. 
+then include bird in X(t), lecture 16
+
+another way: bird is not in X(t)
+
+$$B_{t+dt} = \alpha B^*_{t+dt} - \beta(R_t) B_{t} F_{t}$$
+
+just do the fox equation, no bird equation
 
 ```{r}
-install.packages("pomp")
-library(pomp)
+# compute the average temperature and preciptation. 
+data_frame <- df[,c(1,3,6,7,8:17)]
+data_frame$Temp <- rowMeans(data_frame[,c(5:10)], na.rm = TRUE) 
+data_frame$Precip <- rowMeans(data_frame[,c(11:14)], na.rm = TRUE) 
+data_frame <- data_frame[,c(1,2,3,4,15,16)]
+names(data_frame) <- c('year', 'CUPE', 'R', 'NAO', 'Temp', 'Precip')
+data_frame$R <- ifelse(data_frame$R == "yes", 1, 0)
+data_frame <- transform(data_frame, year = as.numeric(year))  # Ensure 'year' is numeric
+data_frame <- data_frame[c(3:142), ]
+
 ```
 
-To create a POMP for CPUE, I need to define the state process that reflects the underlying dynamics of the species population and the observation process that relates the state to the CPUE measurements.
+Euler Approximation of the predator population at time $t+ \delta t$...
+
+$$F_{t+\delta t} = F_t + \delta(t) F_t B_t \delta t - \theta F_t \delta t$$
+$$F_{t+\delta t} = F_t + (\phi^R \delta_1 F_t B_t - \theta F_t) \delta t$$
+
+$\delta(t) F_t B_t \delta t$ is the expected number of new predators arising from the Poisson process with rate $\delta(t) F_t B_t$;
+$- \theta F_t \delta t$ is the deterministic decrease in predators due to natural death over the time step $\delta t$
+
+$$B_{t+ \delta t} = B_t + \alpha B_t \delta t - \beta(R_t) B_t F_t\delta t$$
+$$B_{t+ \delta t} = B_t + (\alpha B_t - \gamma^R \beta B_t F_t) \delta t$$
+
+```{r}
+meas <- data_frame %>% select(year, reports=CUPE, R) %>% as.data.frame()
+
+sir_step <- function (Fox, B, psi, gamma, alpha, delta_1, beta_1, delta.t,theta, ...) {
+  dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
+  dB = rpois(1, alpha * B) - rpois(1, pow(gamma, R) * beta_1 * B * Fox)
+
+  Fox <- Fox + dF - theta * Fox * delta.t
+  B <- B + dB
+  c(Fox = Fox, B = B)
+}
+
+sir_rinit <- function (...) {
+  c(Fox = 90, B = 90)
+}
+
+meas |>
+  pomp(times="year",t0=meas$year[1]-1,
+    rprocess=euler(sir_step,delta.t=1),
+    rinit=sir_rinit
+  ) -> measSIR
+
+# 
+sir_step<- Csnippet("
+  double dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
+  double dB = rpois(1, alpha * B) - rpois(1, pow(gamma, R) * beta_1 * B * Fox)
+
+  Fox += dF -theta * Fox * delta.t;  
+  B += dB;  
+")
+
+# dmeasure: Define the measurement model
+sir_dmeas <- Csnippet("
+  lik = dnbinom_mu(reports, size = rho * B, prob = exp(-sigma), give_log);
+")
+
+# rmeasure: 
+sir_rmeas <- Csnippet("
+  reports = rnbinom_mu(rho * B, prob = exp(-sigma));
+")
+## size =  number of successes until the experiment is stopped; success rate / hunting rate:
+
+# initialize:
+sir_rinit <- Csnippet("
+  Fox = 90;
+  B = 90;
+")
+
+# Define parameters and initial states
+params <- c(alpha = 0.05, beta_1 = 0.02, gamma = 2, delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.5, sigma = 1)
+
+# create a pomp object
+measSIR |>
+  pomp(rprocess=euler(sir_step,delta.t=1),
+    rinit=sir_rinit,
+    rmeasure=sir_rmeas,
+    dmeasure=sir_dmeas,
+    statenames=c("Fox","B"),
+    paramnames=c("alpha", "gamma", "delta_1", 'beta_1', "psi", "theta", "rho" ,'sigma'),
+    obsnames = c('R')
+  ) -> measSIR
+
+```
+
+```{r}
+meas <- data_frame %>% select(year, reports=CUPE, R) %>% as.data.frame()
+
+sir_step <- function (Fox, psi, delta_1, delta.t, R, B, theta, ...) {
+  dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
+  Fox <- Fox + dF - theta * Fox * delta.t
+  Fox = Fox
+
+}
+
+sir_rinit <- function (...) {
+  Fox = 90
+}
+
+meas |>
+  pomp(times="year",t0=meas$year[1]-1,
+    rprocess=euler(sir_step,delta.t=1),
+    rinit=sir_rinit
+  ) -> measSIR
+
+# 
+sir_step<- Csnippet("
+  double dF = rpois(1, pow(psi, R) * delta_1 * Fox * B);
+  Fox += Fox + dF - theta * Fox * delta.t;
+")
+
+# dmeasure: Define the measurement model
+sir_dmeas <- Csnippet("
+  lik = dnbinom_mu(reports, size = rho * B, prob = exp(-sigma), give_log);
+")
+
+# rmeasure: 
+sir_rmeas <- Csnippet("
+  reports = rnbinom_mu(rho * B, prob = exp(-sigma));
+")
+## size =  number of successes until the experiment is stopped; success rate / hunting rate:
+
+# initialize: number of red foxes at 1983
+sir_rinit <- Csnippet("
+  Fox = 90;
+")
+
+# Define parameters and initial states
+params <- c(delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.5, sigma = 1)
+
+# create a pomp object
+measSIR |>
+  pomp(rprocess=euler(sir_step,delta.t=1),
+    rinit=sir_rinit,
+    rmeasure=sir_rmeas,
+    dmeasure=sir_dmeas,
+    statenames=c("Fox",'B'),
+    paramnames=c("delta_1", "psi", "theta", "rho" ,'sigma'),
+    obsnames = c('R')
+  ) -> measSIR
+
+```

### 2024-04-22 draft.Rmd
changes to draft

--- 2024-04-19 draft.Rmd
@@
 Csnippet("
-
-") -> rproc
-
-Csnippet("
   count = dnbinom(rho*B, sigma)
@@
 
-mod <- pomp(rprocess = euler(step, delta.t =),
+mod <- pomp(rprocess = euler(step, delta.t = 1/52),
             times = "year", t0 = 1872,
             rinit = rinit,
-            rprocess = eurler(step, delta.t=),
+            rprocess = euler(step, delta.t=),
             rmeasure = rmeas, 
@@
 ```
+
+- `rinit`: simulator for the initial-state distiburution, i.e., the distribution of the latent state at time $t_0$
 

### 2024-04-22 draft.Rmd

--- 2024-04-19 draft.Rmd
@@
 - **year**: Self explanatory
-- **CPUE**: Catch per unit effort, expressed as a number of birds shot per hunter per year.
-- **log.CPUE**: Logarithm transformation of CPUE
+- **log.CPUE**: Logarithm transformation of CPUE, catch per unit effort, expressed as a number of birds shot per hunter per year.
 - **peak_rodent_year**: Peak rodent year is scored as “yes”, otherwise “no”. Occurrence of peak
@@
 (Myrberget 1982a), 1971-1979 (Christiansen 1983), 1981-1985 (Frafjord 1988), 1986- 1989
-(Selås et al. 2013), and 1990-2013 (Framstad 2017). Note that 1980 is missing as we were
-not able to find an estimate for this year.
-- **NAO**: The mean North Atlantic Oscillation index for the months May through
-July.
+(Selås et al. 2013), and 1990-2013 (Framstad 2017).
 
-```{r libraries, echo=FALSE, message=FALSE}
+```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
 library(readxl)
@@
 data |>
-  rename(peak_rodent_year = "peak rodent year") |>
-  rename(NAO = "NAO May_Jun_Jul")-> data
+  rename(R = "peak rodent year") |>
+  rename(NAO = "NAO May_Jun_Jul") |>
+  rename(logCPUE = "log.CPUE") |>
+  mutate(R = ifelse(R == "no", 0, 1)) -> data
 
-knitr::kable(data[1:5, 1:7], caption = "Data")
+subset <- c("year", "logCPUE", "R")
+
+knitr::kable(data[subset], caption = "Data")
 ```
@@
   geom_rect(
-    data = subset(data, peak_rodent_year == 'yes'),
+    data = subset(data, peak_rodent_year == 1),
             aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
@@
 This simple model will serve as a reference against which the performance of the more advanced POMP model is measured.
-Essentially, the ARMA model will serve as a yardstick for managing expectations.
+In theory, the ARMA model will serve as a yardstick for managing expectations.
 Of the models shown in (table #), we select the one with the Akaike’s information criterion (AIC).
@@
 
-```{r, echo=FALSE}
-# This code snippet is to add a new column that mutates peak_rodent_year into 
-# binary values for the POMP model.
-data |>
-  mutate(rodent_binary = ifelse(peak_rodent_year == 'yes', 1, 0)) -> data
-```
-
 Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models "consist of an unobserved Markov state process, connected to the data via an explicit model of the observation process." The former is referred to as the *latent process model* and the latter as the *measurement model.* A simple visual representation of the process is depicted in figure [].
@@
 
-The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator and alternative prey (rodent) count.
+The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator (fox), preferred prey(rodent) and alternative prey (bird) count.
 
@@
 
-- $R(t)=$ peak rodent year, from data
+- $Z(t)= R(t)$ peak rodent yea
+
+$$Y_n = N$$
 
@@
 
+```{r}
+statenames <- c("logF", "logB")
+paramnames <- c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
+                "logF_0", "logB_0", "logRho", "sigma_obs")
+obs_names <- c("logCPUE")
+covarnames <- c("R")
+# logF_0 = 0 normalizing to assume unit initial population
+# sigmaF roughly the noise in a year
+param_guess <- c(
+  logF_0 = 0, 
+  logB_0 = 0,
+  a = 1,
+  b = 2, 
+  c = 1,
+  alpha = 1, 
+  Beta = 1,
+  gamma = 0.5,
+  sigmaF = 0.1, 
+  sigmaB = 0.1, 
+  logRho = 3, 
+  sigma_obs = 0.1
+)
+```
+
+```{r}
+bird <- data[subset]
+head(bird)
+```
+
 ```{r, eval=FALSE}
 Csnippet("
-  double dNf  = rbinom(F, 1-exp);
-  double dF_p = rbinom(F, 1-exp(rho_t * beta_t * dt));
-  double dF_m = rbinom(F, 1-exp(theta*dt));
+  double dwW, dwB;
 
-  B = 
-  F += dF_p - dF_m;
+  dwF = rgammawn(sigmaF, dt);
+  dwB = rgammawn(sigmaB, dt);
 
+  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
+  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
 ") -> step
@@
 Csnippet("
-  B = 
-  F = 
+  logF = logF_0;
+  logB = logB_0;
 ") -> rinit
@@
 Csnippet("
-
-") -> rproc
+  log.CPUE = rnorm(logB + logrho, signma_obs)
+") -> rmeas
 
 Csnippet("
-  count = dnbinom(rho*B, sigma)
-") -> rmeas
-
-Cnsippet("
-
+  lik = dnorm(logCPUE, lobB + logrho, sigma_obs.give_log)
 ") -> dmeas
 
-mod <- pomp(rprocess = euler(step, delta.t =),
-            times = "year", t0 = 1872,
-            rinit = rinit,
-            rprocess = eurler(step, delta.t=),
-            rmeasure = rmeas, 
-            dmeasure = , 
-            paramnames = c("rho", "theta", "delta"),
+bird |>
+  pomp(times="year", t0=1872,
+       rprocess=euler(step, delta.t=1/52),
+       rinit=rinit,
+       rmeasure=rmeas,
+       dmeasure=dmeas,
+       statenames=statenames,
+       paramnames=paramnames,
+       obs_names <- c("logCPUE"),
+       covarnames <- c("R")
+    ) -> mod
 
-)
 ```
+
+- `rinit`: simulator for the initial-state distribution, i.e., the distribution of the latent state at time $t_0$
 

### 2024-04-22 workspace/heyuan.Rmd
Add files via upload

--- 2024-04-16 workspace/heyuan.Rmd
@@
 
-## POMP 1
+## POMP 3
+
+Y(t): bird count, $y_t = NB(mean = \rho B_t, \sigma)$, $B_t$ is bird population size at time t
+X(t): unobserved state process, red foxes, Bird, Peak Rodent Year, temperature, precipitation, NAO index
+
+Reference: Lotka-Volterra predator-prey model for the state process: $$X_{t+dt} = \beta X_t - \mu X_t$$
+
+natural born - number of birds eaten by foxes: 
+$$B_{t+dt} = \alpha B_t - \beta(R_t) B_{t}F_{t}$$
+
+where, $$\beta(R_t) = \beta if R(t)=0$$ or 
+$$ \beta(R_t) = \gamma \beta, \gamma > 1$$ , rodent number affect the rate of fox eating birds, 
+$\beta(R_t)$ can be re-written as $$\beta(R_t) =  \gamma^R \beta$$, where R = 0 or 1
+
+Growth Rate - Death Rate of Foxes: 
+$$ F_{t+dt} = \delta(t) F_t B^*_t - \theta(t) F_t$$
+
+where, $$\delta(t) = \delta_1 if R(t) = 0, or, \phi \delta_1$$ , $\phi<1$ and whether it is in rodent peak year affect the number of birds and the number of foxes.
+
+$\delta(t)$ can be re-written as $$\delta(t) = \phi^R \delta_1$$, where R = 0 or 1
+
+$\theta(t)$ could include the temperature, NAO about how does it affect the death rate of Foxes.
+
+first, let dt=1 year
+then, dt = less than 1 year, maybe 1 month, bird keeps as 100, foxes=10, 15, 17, 15 for each dt until reaching 1 year. 
+then include bird in X(t), lecture 16
+
+another way: bird is not in X(t)
+
+$$B_{t+dt} = \alpha B^*_{t+dt} - \beta(R_t) B_{t} F_{t}$$
+
+just do the fox equation, no bird equation
 
 ```{r}
-install.packages("pomp")
-library(pomp)
+# compute the average temperature and preciptation. 
+data_frame <- df[,c(1,3,6,7,8:17)]
+data_frame$Temp <- rowMeans(data_frame[,c(5:10)], na.rm = TRUE) 
+data_frame$Precip <- rowMeans(data_frame[,c(11:14)], na.rm = TRUE) 
+data_frame <- data_frame[,c(1,2,3,4,15,16)]
+names(data_frame) <- c('year', 'CUPE', 'R', 'NAO', 'Temp', 'Precip')
+data_frame$R <- ifelse(data_frame$R == "yes", 1, 0)
+data_frame <- transform(data_frame, year = as.numeric(year))  # Ensure 'year' is numeric
+
 ```
 
-To create a POMP for CPUE, I need to define the state process that reflects the underlying dynamics of the species population and the observation process that relates the state to the CPUE measurements.
+Euler Approximation of the predator population at time $t+ \delta t$...
+
+$$F_{t+\delta t} = F_t + \delta(t) F_t B_t \delta t - \theta F_t \delta t$$
+$$F_{t+\delta t} = F_t + (\phi^R \delta_1 F_t B_t - \theta F_t) \delta t$$
+
+$\delta(t) F_t B_t \delta t$ is the expected number of new predators arising from the Poisson process with rate $\delta(t) F_t B_t$;
+$- \theta F_t \delta t$ is the deterministic decrease in predators due to natural death over the time step $\delta t$
+
+$$B_{t+ \delta t} = B_t + \alpha B_t \delta t - \beta(R_t) B_t F_t\delta t$$
+$$B_{t+ \delta t} = B_t + (\alpha B_t - \gamma^R \beta B_t F_t) \delta t$$
+
+```{r, warning=FALSE}
+meas <- dplyr::select(data_frame, year, reports=CUPE, R)
+
+sir_step<- Csnippet("
+  double rateBin, rateBout, rateFin, rateFout;
+
+  rateBin = alpha * B * dt;
+  rateBout = pow(gamma, R) * beta_1 * B * Fox * dt;
+  rateFin = pow(psi, R) * delta_1 * Fox * B * dt;
+  rateFout = theta * Fox * dt;
+
+  // double dF = rbinom(Fox, 1 - exp(-rateF));
+  double dFin = rbinom(Fox, 1 - exp(-rateFin));
+  double dFout = rbinom(Fox, 1 - exp(-rateFout));
+
+  double dBin = rbinom(B, 1 - exp(-rateBin));
+  double dBout = rbinom(B, 1 - exp(-rateBout));
+
+  Fox += dFin - dFout;  
+  B += dBin - dBout;
+  B = fmax(B, 0.0); 
+  Fox = fmax(Fox, 0.0);
+  // Rprintf(\"B = %f\\n\\n \", B);
+
+")
+
+# dmeasure: Define the measurement model
+sir_dmeas <- Csnippet("
+  if (give_log) {
+    if (B == 0 & reports != 0) {
+      lik = -100000;
+    } else {
+      // Rprintf(\"rho * B = %f\\n \", rho * B);
+      // Rprintf(\"reports = %f\\n\\n \", reports);
+      lik = dnbinom_mu(reports, k, rho * B, give_log);
+    }
+  } else {
+    lik = dnbinom_mu(reports, k, rho * B, give_log);
+  }
+")
+
+# rmeasure: 
+sir_rmeas <- Csnippet("
+  reports = rnbinom_mu(k, rho * B);
+")
+## size =  number of successes until the experiment is stopped; success rate / hunting rate:
+
+# initialize:
+sir_rinit <- Csnippet("
+  Fox = nearbyint(F_0);
+  B = nearbyint(B_0);
+")
+
+# Define parameters and initial states
+params <- c(alpha = 0.5, beta_1 = 0.0001, gamma = 1, delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.01, k = 20, F_0 = 90, B_0 = 1000)
+
+my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')
+
+# create a pomp object
+pomp(data = meas |> dplyr::select(-R) |> as.data.frame(),
+     rprocess=euler(sir_step,delta.t=1/12), times='year', t0=1871,
+    rinit=sir_rinit,
+    rmeasure=sir_rmeas,
+    dmeasure=sir_dmeas,
+    statenames=c("Fox","B"),
+    paramnames=c("alpha", "gamma", "delta_1", 'beta_1', "psi", "theta", "rho" ,'k', 'F_0', 'B_0'),
+    covar = my_covars,
+    partrans = parameter_trans(log = c('alpha', 'gamma', 'delta_1', 'beta_1', 'psi', 'theta', 'rho', 'k', 'F_0', 'B_0'))
+  ) -> measSIR
+
+measSIR |>
+  pfilter(Np=5, params = params) -> pf
+
+mif2(
+  measSIR, 
+  Nmif = 10, 
+  Np = 2000,
+  cooling.fraction.50 = 0.5, 
+  rw.sd = rw_sd(
+    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
+  ),
+  params = params
+) -> mif_out
+
+```
+
+```{r}
+statenames <- c('logF', 'logB')
+paramnames <- c('a', 'b', 'c', 'gamma', 'alpha', 'Beta', 'sigmaF', 'sigmaB',
+                'logF_0', 'logB_0', 'logRho', 'sigma_obs')
+obs_names <- c('logCPUE')
+# logF_0 = 0 normalizing to assume unit initial population
+# sigmaF roughly the noise in a year
+param_guess <- c(
+  logF_0 = 0,
+  logB_0 = 0,
+  a = 1,
+  b = 2,
+  c = 1,
+  alpha = 1,
+  Beta = 1,
+  gamma = 0.5,
+  sigmaF = 0.1,
+  sigmaB = 0.1,
+  logRho = 3,
+  sigma_obs = 0.1
+)
+Csnippet('
+  double dwB, dwF;
+  dwF = rgammawn(sigmaF, dt);
+  dwB = rgammawn(sigmaB, dt);
+  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
+  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
+') -> rproc
+Csnippet('
+  logF = logF_0;
+  logB = logB_0;
+') -> rinit
+Csnippet('
+  log.CPUE = rnorm(logB + logrho, signma_obs);
+') -> rmeas
+Csnippet('
+  lik = dnorm(logCPUE, logB + logrho, sigma_obs, give_log);
+') -> dmeas
+
+my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')
+
+pomp(data = data_frame |> dplyr::select(-R),rprocess=euler(rproc,delta.t=1/12),
+     times = 'year', t0 = 1871,
+    rinit=rinit,
+    # rmeasure=rmeas,
+    # dmeasure=dmeas,
+    statenames=statenames,
+    paramnames=paramnames,
+    covar = my_covars
+  ) -> measSIR
+
+```

### 2024-04-23 draft.Rmd
n

@@
 
-Studying ecology, especially animal population dynamics, is essential for understanding the complex interactions between species and their environment, guiding conservation efforts, and ensuring the sustainable use of natural resources for future generations..
-By modeling population dynamics, researchers and scientists can understand how animals respond to environmental changes and human activities.
-This helps conservation efforts by identifying threatened species and develop strategies to protect biodiversity.
-Moreover, animal populations can serve as reservoirs for infectious diseases that can affect both human and animal health.
-Having a robust understanding of population movement can help predict disease outbreaks, understand transmission patterns, and improve disease control and prevention.
+Studying population dynamics is essential for understanding the complex interactions between species and their environment.
+By modeling population dynamics, researchers can gain an understanding of how animals respond to environmental changes and activities.
+This understanding is vital for conservation efforts as it aids in identifying endangered species and devising strategies to safeguard biodiversity.
+Furthermore, animal populations can act as reservoirs for infectious diseases, impacting both human and animal health.
+By comprehensively understanding population movements, scientists can predict outbreaks, discern transmission patterns, and enhance disease control and prevention measures.
 
-This report extends previous research investigating the long-term population decline of willow ptarmigan *Lagopus lagopus*.
-
-- The alternative prey hypothesis predicts that the interaction between generalist predators and their main prey is a major driver of population dynamics of alternative prey species.
-
-- Grouse species are of the most studied
-
-- In this report, we analyze 142 years of data to investigate whether the population growth of a declining alternative prey species, the willow ptarmigan, is related to an interaction between red foxes and their main prey, small rodents.
-
-Two types of questions we can ask.
-
-- Scientific goals: Which explanations are most favored by the data? Which kinds of data are most informative?
-- Applied goals: how to design ecological or epidemiological intervention? How to make accurate forecasts?
-
-This report seeks to answer two scientific questions.
-First, between alternative prey hypothesis and the role of climate change, which explanation is most favored by the data?
-Secondly, which kind of data is more informative?
+This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan Lapagog lapagos in Northeastern Scandinavia ().
+Although not in immediate danger, the willow ptarmigan faces threats from habitat loss, climate change, and hunting and trapping.
+Hjeljord and Loe postulate that, along with climate, a long-term dampening of the amplitude in small rodents and an increase in red fox numbers, have prevented willow ptarmigan populations from reaching their peaks seen a hundred years ago ().
+Their analysis implements linear models with a count proxy as a function of time to estimate linear change and wavelet analysis to detect cyclic periods.
+This report’s aim is to further Hjelford and Loe’s work by capturing the stochastic population dynamics and the role of alternative prey using the partially observed markov process (POMP) framework
 
@@
 
-In ecological studies, researching predator-prey dynamics, especially within the context of the alternative prey hypothesis, offers profound understanding into ecosystem complexities.
-Focusing on ptarmigans in Norway, a number of studies have attempted to model the interplay between ptarmigans as prey and their predators, including gyrfalcons and foxes.
-Notably, research by Aanes et al.
-(2002) highlighted the significant impact of predation pressure on ptarmigan populations, emphasizing the need to understand predator-prey dynamics for effective conservation strategies.
-
-## Objectives
-
-Our research objectives are as follow:
-
-- 
-- There has been a lack in application of the Partially Observed Markov Process (POMP) framework to ecological research questions.
+Surprisingly, grouse, the family of birds which willow ptarmigan belong to, are of the most studied bird species in the world.
+Thus, an abundance of scientific research has been carried out to shed light on the decline in its population.
+@pedersen found that intensified forestry practices have significantly reduced suitable breeding habitat.
+Additionally, agricultural expansion has led to the conversion of tundra and heathland habitats, further diminishing ptarmigan populations [@henden].
+Climate change exacerbates these effects by altering snow cover patterns, impacting the ptarmigans' ability to camouflage and increasing predation risk.
+Research by @tottrup highlighted that warmer temperatures and reduced snow cover are leading to increased predation pressure on ptarmigans in Scandinavia.
+Furthermore, hunting pressure has also played a role in the decline of ptarmigan populations.
+Studies by @hebbelstrup and @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.
+Breisjøberget el al., found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis.
 
@@
 We briefly introduce and describe our data set and variables of interest.
-Following that, a short exploration of data trend, correlation, and association is is given.
+Followed by a short exploration of trend, correlation, and association.
 Next, we build a baseline ARMA process model to serve as a measurement for our more complex proposals.

### 2024-04-23 draft.Rmd
uploaded working model

@@
 csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
+bibliography: references.bib 
 output: 
@@
 
-This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan Lapagog lapagos in Northeastern Scandinavia ().
+This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan *Lapagos lapagos* in Northeastern Scandinavia ().
 Although not in immediate danger, the willow ptarmigan faces threats from habitat loss, climate change, and hunting and trapping.
@@
 Thus, an abundance of scientific research has been carried out to shed light on the decline in its population.
-@pedersen found that intensified forestry practices have significantly reduced suitable breeding habitat.
+Pedersen el al., @pedersen found that intensified forestry practices have significantly reduced suitable breeding habitat.
 Additionally, agricultural expansion has led to the conversion of tundra and heathland habitats, further diminishing ptarmigan populations [@henden].
 Climate change exacerbates these effects by altering snow cover patterns, impacting the ptarmigans' ability to camouflage and increasing predation risk.
-Research by @tottrup highlighted that warmer temperatures and reduced snow cover are leading to increased predation pressure on ptarmigans in Scandinavia.
+Research by Tottrup et al., @tottrup highlighted that warmer temperatures and reduced snow cover are leading to increased predation pressure on ptarmigans in Scandinavia.
 Furthermore, hunting pressure has also played a role in the decline of ptarmigan populations.
-Studies by @hebbelstrup and @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.
+Studies by Hebbelstrup @hebbelstrup and Storass @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.
 Breisjøberget el al., found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis.
+Despite there being a plethora of work conducted in this area, none have applied
 
@@
 library(cowplot)
+library(doParallel)
 library(pomp)
@@
 ```{r data_table, echo=FALSE}
-data <- read_excel("./data/bird_data.xlsx")
-data |>
+df <- read_excel("./data/bird_data.xlsx")
+df |>
   rename(R = "peak rodent year") |>
@@
 
-knitr::kable(data[subset], caption = "Data")
+knitr::kable(head(data[subset], 10), caption = "Data")
 ```
@@
 
-p2 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
+p2 <- ggplot(data, aes(x=year, y=logCPUE)) + 
   geom_line(color="steelblue") + 
@@
 
-p3 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
+p3 <- ggplot(data, aes(x=year, y=logCPUE)) + 
   geom_line(color="steelblue") + 
@@
   geom_rect(
-    data = subset(data, peak_rodent_year == 1),
+    data = subset(data, R == 1),
             aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
@@
 ```{r, boxplot, echo=FALSE}
-ggplot(data, aes(x = peak_rodent_year, y = log.CPUE)) +
-
+ggplot(data, aes(x = factor(R), y = logCPUE, group = factor(R))) + 
   geom_boxplot() + 
   theme_light() + 
-  xlab("Peak Rodent Year")
+  xlab("Peak Rodent Year") +
+  scale_x_discrete(labels = c("no" = "0", "yes" = "1"))
 ```
@@
 ```{r ACF, echo=FALSE}
-acf(data$log.CPUE, lag = 100, main = "ACF log.CPUE")
+acf(data$logCPUE, lag = 100, main = "ACF log.CPUE")
 ```
@@
 ```{r PACF, echo=FALSE}
-pacf(data$log.CPUE, lag = 100, main = "PACF log.CPUE")
+pacf(data$logCPUE, lag = 100, main = "PACF log.CPUE")
 ```
@@
 
-bird_aic_table <- aic_table(data$log.CPUE,4,5)
+bird_aic_table <- aic_table(data$logCPUE,4,5)
 require(knitr)
@@
 
+```{r, echo=FALSE, eval=FALSE}
+df |>
+  rename(R = "peak rodent year") |>
+  rename(NAO = "NAO May_Jun_Jul") |>
+  rename(logCPUE = "log.CPUE") |>
+  mutate(R = ifelse(R == "no", 0, 1))  |>
+  rename(reports = "CPUE") |>
+  select(c("year", "reports", "R")) -> bird
+```
+
+```{r echo=FALSE, eval=FALSE}
+birdMod |>
+  pfilter(Np=5, params = params) -> pf
+
+mif2(
+  measSIR, 
+  Nmif = 10, 
+  Np = 2000,
+  cooling.fraction.50 = 0.5, 
+  rw.sd = rw_sd(
+    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
+  ),
+  params = params
+) -> mif_out
+```
+
+### Local Search
+
 ```{r}
@@
 obs_names <- c("logCPUE")
-covarnames <- c("R")
 # logF_0 = 0 normalizing to assume unit initial population
 # sigmaF roughly the noise in a year
+```
+
+```{r}
+bird <- data[subset]
+head(bird)
+```
+
+```{r}
+Csnippet("
+  double dwF, dwB;
+
+  dwF = rgammawn(sigmaF, dt);
+  dwB = rgammawn(sigmaB, dt);
+
+  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
+  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
+") -> step
+
+Csnippet("
+  logF = logF_0;
+  logB = logB_0;
+") -> rinit
+
+Csnippet("
+  logCPUE = rnorm(logB - logRho, sigma_obs);
+") -> rmeas
+
+Csnippet("
+  lik = dnorm(logCPUE, logB - logRho, sigma_obs, give_log);
+") -> dmeas
+```
+
+```{r}
+bird1 <- bird |> 
+  dplyr::select(-R)
+
+bird1 |>
+  pomp(
+    times="year", 
+    t0=1871,
+    rprocess=euler(step, delta.t=1/52),
+    rinit=rinit,
+    rmeasure=rmeas,
+    dmeasure=dmeas,
+    statenames=statenames,
+    paramnames=paramnames,
+    obs_names <- c("logCPUE"),
+    covar = covariate_table(
+      bird |> dplyr::select(year, R) |> 
+        dplyr::bind_rows(c("year" = 1871, R = 0)) |> 
+        dplyr::arrange(year), 
+      times = 'year',
+      order = "constant"
+    ),
+    partrans = parameter_trans(
+      log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
+              'c', 'logF_0', 'logB_0', 'logRho')
+    )
+    # covarnames <- c("R")
+  ) -> mod
+```
+
+```{r}
 param_guess <- c(
-  logF_0 = 0, 
-  logB_0 = 0,
+  logF_0 = 1, 
+  logB_0 = 1,
   a = 1,
@@
 ```{r}
-bird <- data[subset]
-head(bird)
+mif2(
+  mod, 
+  Np = 1000, 
+  Nmif = 50, 
+  rw.sd = rw_sd(
+    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
+    a = 0.02, b = 0.02, c = 0.02, 
+    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
+    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
+  ),
+  cooling.fraction.50 = 0.5,
+  params = param_guess
+) -> mif2_out
 ```
 
-```{r, eval=FALSE}
-Csnippet("
-  double dwW, dwB;
-
-  dwF = rgammawn(sigmaF, dt);
-  dwB = rgammawn(sigmaB, dt);
-
-  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
-  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
-") -> step
-
-Csnippet("
-  logF = logF_0;
-  logB = logB_0;
-") -> rinit
-
-Csnippet("
-  log.CPUE = rnorm(logB + logrho, signma_obs)
-") -> rmeas
-
-Csnippet("
-  lik = dnorm(logCPUE, lobB + logrho, sigma_obs.give_log)
-") -> dmeas
-
-bird |>
-  pomp(times="year", t0=1872,
-       rprocess=euler(step, delta.t=1/52),
-       rinit=rinit,
-       rmeasure=rmeas,
-       dmeasure=dmeas,
-       statenames=statenames,
-       paramnames=paramnames,
-       obs_names <- c("logCPUE"),
-       covarnames <- c("R")
-    ) -> mod
-
+```{r}
+freeze(
+  runif_design(
+    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
+              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
+              sigmaF = 0, sigmaB = 0.02, logRho = 0)
+    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
+              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
+              sigmaF = 0, sigmaB = 0.02, logRho = 0)
+  ),
+  seed = 420
+) -> guesses
 ```
 
-- `rinit`: simulator for the initial-state distribution, i.e., the distribution of the latent state at time $t_0$
+### Global Search
 

### 2024-04-23 workspace/sizhuang.Rmd
pomp equations

--- 2024-04-07 draft.Rmd
@@
----
-title: "Draft"
-author: "Pete, Sizhuang, and Heyuan"
-date: "`r Sys.Date()`"
-output: 
-  html_document:
-    theme: cosmo
-    toc: yes
-    toc_float:
-      collapsed: true
----
+```{r}
+library(tidyverse)
+library(tseries)
+#install.packages("ggfortify")
+library(ggfortify)
+library(pomp)
+```
 
-![Female Willow Ptarmigan captured by [Bryce W. Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](./assets/ptarmigan.jpg)
+#### read in the data & data description
+```{r}
+data_path <- '//Users/sizhuang/Desktop/GitHubRepos/pomp_final/data/data.txt'
+df <- read.table(data_path, header = TRUE, sep = "\t", na.strings = "NA", dec = ".", strip.white = TRUE)
+head(df)
 
-# Introduction
+```
 
-## Related Work
+142 years (1872-2013) of willow ptarmigan hunting records in southeastern Norway
 
-## Data
+CPUE: the number of birds shot per hunter per day, as the primary metric for assessing population changes.
+
+NAO: A positive NAO index favors mild winters in the U.S. East. Negative values favor stronger cold-air outbreaks and increased storminess in the eastern U.S.
+
+temp.PREINC: temperature, pre-incubation, 5/1 - 6/2
+temp.INC: temperature, incubation, 6/3 - 6/24
+temp.BROOD: temperature, brooding & rearing, 6/25 - 7/15
+temp.HATCH: temperature, mean hatching date, 7/21 - 7/27
+prefix prec.: precipitation
+
+opening data hunt(julian day): the starting day of hunting season is named as Julian Day, which is varied across 142 years.
+It is mainly in mid to late August up to the 1920's and 10th/15th September in later years.
 
@@
 
-```{r libraries}
-library(readxl)
-library(ggplot2)
-library(knitr)
+#### Trend Analysis
+Temporal Patterns: CPUE showed significant declines from early records to recent years, with notable shifts in population cycles and their amplitude.
+
+Impact of Alternative Prey: The availability of rodents, which serve as alternative prey for predators that also hunt ptarmigan, was strongly linked to fluctuations in ptarmigan numbers.
+Years with high rodent populations typically saw higher CPUE for ptarmigan, suggesting lower predatory pressure on them.
+
+#### Correlation
+For the ACF plot, the initial 40 lags show a strong correlation with past values, diminishing quickly after first 5 lags.
+It suggests $q = 5$, $q = 40$ or any value smaller than 40 appropriate for MA(q) component.
+For the PACF plot, partial correlation dies out quickly after 5 lags.
+It would be appropriate to select $p = 0$ or $p = 5$ for AR(p) component.
+
+Given the long-range significance in the ACF, I may be dealing with a non-stationary series that needs differencing.
+Also this plot with a flip to negative autocorrelations and a cyclical pattern that declines over time is a characteristic sign of seasonality in my data.
+However, STL or classical decomposition fails because time series has no or less than 2 periods, indicating the non-existence of seasonality.
+So the repeated short-term cycle in 5 lags could be explained by the partial correlation.
+
+```{r, error=TRUE}
+start_year <- min(df$year)
+end_year <- max(df$year)
+frequency <- 1  
+ts_cpue <- ts(df$log.CPUE, start=start_year, end=end_year, frequency=frequency)
+
+# Apply STL decomposition
+ts_components <- stl(ts_cpue, s.window = "periodic")
+plot(ts_components)
 ```
 
-```{r data_table}
-data <- read_excel("./data/bird_data.xlsx")
+#### Stationary Check
+ARMA(p,q) is performed only if time series data is stationary. So, I will do the stationarity check by ADF.
 
-knitr::kable(data[1:5, 1:5], caption = "Data")
+```{r}
+# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
+kpss_result <- kpss.test(ts_cpue, null = "Level")
+print(kpss_result)
+```
+
+From ADF test, p-value is estimated as 0.01, smaller than 0.05. We have enough evidence to reject the null hypothesis that the time series is non-stationary.
+
+KPSS with p-value = 0.01 suggests that the log CPUE times series data is non-stationary. This means I can't fit the data with ARMA(p,q) without differencing to achieve stationarity.
+
+#### Differencing
+```{r}
+# Differencing the time series once
+ts_diff <- diff(ts_cpue, differences = 1)
+print(kpss.test(ts_diff,null = "Level"))
+```
+
+KPSS with p-value = 0.1 suggests that the first-order differenced times series data is stationary. This means I can fit this differenced data with ARMA(p,q).
+
+# Methods
+
+## ARMA
+From ADF test, p-value is estimated as 0.01, smaller than 0.05.
+We have enough evidence to reject the null hypothesis that the time series is non-stationary.
+KPSS also support the same conclusion that this times series data is stationary.
+Thus, we could conclude that CPUE on log scale is stationary.
+This means I can fit the data with ARMA(p,q) without the need to difference or other ways of data transformation to achieve stationarity.
+
+```{r}
+
+aic_table <- function(data,P,Q){
+  table <- matrix(NA,(P+1),(Q+1))
+  for(p in 0:P) {
+    for(q in 0:Q) {
+      table[p+1,q+1] <- arima(data,order=c(p,0,q))$aic
+    }
+  }
+  dimnames(table) <- list(paste("AR",0:P, sep=""),
+    paste("MA",0:Q,sep=""))
+  table
+}
+
+bird_aic_table <- aic_table(ts_diff,5,10)
+require(knitr)
+kable(bird_aic_table,digits=2)
+
+```
+
+The smaller AIC, the better the model to fit the data. Thus, ARMA(0,5) with the lowest AIC value of 204.48 is the best model.
+
+```{r}
+aic_table <- function(data,P,Q){
+  table <- matrix(NA,(P+1),(Q+1))
+  for(p in 0:P) {
+    for(q in 0:Q) {
+      table[p+1,q+1] <- arima(data,order=c(p,1,q))$aic
+    }
+  }
+  dimnames(table) <- list(paste("AR",0:P, sep=""),
+    paste("MA",0:Q,sep=""))
+  table
+}
+
+bird_aic_table <- aic_table(ts_cpue,5,10)
+require(knitr)
+kable(bird_aic_table,digits=2)
+```
+
+ARIMA(0,1,5) achieves the same result for original log CPUE data.
+
+ARMA(0,5) Diagnosis
+```{r}
+fit1 <- arima(ts_cpue, order=c(0,0,5))
+residuals <- residuals(fit1)
+
+# ACF check
+acf(residuals, main="ACF of Residuals")
+
+# Normality check with Q-Q plot
+qqnorm(residuals)
+qqline(residuals, col="red")
+
+# Shapiro-Wilk test
+shapiro.test(residuals)
+
+# Root Check
+ma_coefficients <- coef(fit1)[grepl("ma", names(coef(fit1)))]
+MA_roots <- polyroot(c(1, ma_coefficients))
+#invertible <- all(Mod(MA_roots) > 1)
+#print(paste("Is the model invertible? ", invertible))
+
+autoplot(fit1) +
+  ggtitle("ARMA(0, 5)")
+```
+
+For the ACF plot of the residuals, all autocorrelations are within the confidence bounds (indicated by blue dashed lines), which suggests that there is no significant autocorrelation present.
+Additionally, the QQ plot shows that the data points roughly align with the reference line (colored red), supporting the assumption of normality.
+This is further confirmed by a Shapiro-Wilk test for normality, which returns a p-value of 0.62, indicating that the residuals can be considered normally distributed.
+Moreover, the roots of the characteristic polynomial are outside the unit circle, confirming that the MA(5) process is invertible—a critical condition for model adequacy.
+These diagnostics collectively verify that the ARIMA(0,1,5) model is well-specified and adheres to the necessary assumptions for valid time series analysis.
+
+## POMP 3
+
+### Model Overview
+Our model involves two processes: the observed bird count $Y(t)$ and the unobserved state process $X(t)$. The observed bird count at time $t$ is modeled by a negative binomial distribution:
+$$y_t = NB(mean = \rho B_t, \sigma),$$ where $B_t$ is the bird population size at time $t$ and $\rho$ is a scaling factor.
+The unobserved state process $X(t)$ consists several state variables, including:
+* $F_t$: the number of red foxes at time $t$.
+* $R_t$: indicator variable on whether year $r$ is a peak rodent year ($R_t=1$ if year $t$ is a peak rodent year, zero otherwise).
+* Other environmental factors such as temperature, precipitation and the North Atlantic Oscillation (NAO) index.
+
+Reference: Lotka-Volterra predator-prey model for the state process: $$X_{t+dt} = \beta X_t - \mu X_t$$
+
+### Dynamics of Bird and Fox Populations
+The bird population evolves according to the difference equation:
+$$B_{t+dt} = \alpha B_t - \beta(R_t) B_{t}F_{t},$$
+where:
+* $\alpha$ is the natural grow rate of the bird population.
+* $\beta(R_t)$ is the predating rate, modulated by rodents in the following way
+$$\beta(R_t) =  \gamma^{R_t} \beta_1,$$
+where:
+* $\beta_1$ is the base predating rate for non peak rodent years.
+* $\gamma$ is a scaling factor, taking rodent availability into consideration.
+
+The fox population dynamics can be described as:
+$$ F_{t+dt} = \delta(t) F_t B^*_t - \theta(t) F_t$$
+
+where:
++ $\delta(t)$ is the birth rate parameter, modulating by rodents through
+$$\delta(t) = \phi^{R_t} \delta_1,$$
+where $\delta_1$ is the base fox birth rate on non peak rodent years and $\phi$ is a scaling factor for peak rodent years.
+We impose a restriction $\phi<1$ so that in peak rodent years, birth rates of foxes decrease.
++ $B^*_t$: 
++ $\theta(t)$ represents death rates if foxes, which could depend on environmental factors such as temperature, NAO etc.
+
+### Numerical Approximation 
+We use Euler Approximation to approximate the predator (fox) population at time $t+ \Delta t$.
+$$F_{t+\Delta t} = F_t + \delta(t) F_t B_t \Delta t - \theta(t) F_t \Delta t$$
+or equivalently,
+$$F_{t+\Delta t} = F_t + (\phi^{R_t} \delta_1 F_t B_t - \theta(t) F_t) \Delta t$$
+
+where $\delta(t) F_t B_t \delta t$ is the expected number of new predators arising from the Poisson process with rate $\delta(t) F_t B_t$ and $- \theta F_t \delta t$ is the deterministic decrease in predators due to natural death over the time step $\Delta t$.
+
+$$B_{t+ \Delta t} = B_t + \alpha B_t \Delta t - \beta(R_t) B_t F_t\Delta t$$
+or equivalently,
+$$B_{t+ \Delta t} = B_t + (\alpha B_t - \gamma^{R_t} \beta B_t F_t) \Delta t$$
+
+```{r}
+# compute the average temperature and preciptation. 
+data_frame <- df[,c(1,3,6,7,8:17)]
+data_frame$Temp <- rowMeans(data_frame[,c(5:10)], na.rm = TRUE) 
+data_frame$Precip <- rowMeans(data_frame[,c(11:14)], na.rm = TRUE) 
+data_frame <- data_frame[,c(1,2,3,4,15,16)]
+names(data_frame) <- c('year', 'CUPE', 'R', 'NAO', 'Temp', 'Precip')
+data_frame$R <- ifelse(data_frame$R == "yes", 1, 0)
+data_frame <- transform(data_frame, year = as.numeric(year))  # Ensure 'year' is numeric
+
+```
+
+```{r, warning=FALSE}
+meas <- dplyr::select(data_frame, year, reports=CUPE, R)
+
+sir_step<- Csnippet("
+  double rateBin, rateBout, rateFin, rateFout;
+
+  rateBin = alpha * B * dt;
+  rateBout = pow(gamma, R) * beta_1 * B * Fox * dt;
+  rateFin = pow(psi, R) * delta_1 * Fox * B * dt;
+  rateFout = theta * Fox * dt;
+
+  // double dF = rbinom(Fox, 1 - exp(-rateF));
+  double dFin = rbinom(Fox, 1 - exp(-rateFin));
+  double dFout = rbinom(Fox, 1 - exp(-rateFout));
+
+  double dBin = rbinom(B, 1 - exp(-rateBin));
+  double dBout = rbinom(B, 1 - exp(-rateBout));
+
+  Fox += dFin - dFout;  
+  B += dBin - dBout;
+  B = fmax(B, 0.0); 
+  Fox = fmax(Fox, 0.0);
+  // Rprintf(\"B = %f\\n\\n \", B);
+
+")
+
+# dmeasure: Define the measurement model
+sir_dmeas <- Csnippet("
+  if (give_log) {
+    if (B == 0 & reports != 0) {
+      lik = -100000;
+    } else {
+      // Rprintf(\"rho * B = %f\\n \", rho * B);
+      // Rprintf(\"reports = %f\\n\\n \", reports);
+      lik = dnbinom_mu(reports, k, rho * B, give_log);
+    }
+  } else {
+    lik = dnbinom_mu(reports, k, rho * B, give_log);
+  }
+")
+
+# rmeasure: 
+sir_rmeas <- Csnippet("
+  reports = rnbinom_mu(k, rho * B);
+")
+## size =  number of successes until the experiment is stopped; success rate / hunting rate:
+
+# initialize:
+sir_rinit <- Csnippet("
+  Fox = nearbyint(F_0);
+  B = nearbyint(B_0);
+")
+
+# Define parameters and initial states
+params <- c(alpha = 0.5, beta_1 = 0.0001, gamma = 1, delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.01, k = 20, F_0 = 90, B_0 = 1000)
+
+my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')
+
+# create a pomp object
+pomp(data = meas |> dplyr::select(-R) |> as.data.frame(),
+     rprocess=euler(sir_step,delta.t=1/12), times='year', t0=1871,
+    rinit=sir_rinit,
+    rmeasure=sir_rmeas,
+    dmeasure=sir_dmeas,
+    statenames=c("Fox","B"),
+    paramnames=c("alpha", "gamma", "delta_1", 'beta_1', "psi", "theta", "rho" ,'k', 'F_0', 'B_0'),
+    covar = my_covars,
+    partrans = parameter_trans(log = c('alpha', 'gamma', 'delta_1', 'beta_1', 'psi', 'theta', 'rho', 'k', 'F_0', 'B_0'))
+  ) -> measSIR
+
+measSIR |>
+  pfilter(Np=5, params = params) -> pf
+
+mif2(
+  measSIR, 
+  Nmif = 10, 
+  Np = 2000,
+  cooling.fraction.50 = 0.5, 
+  rw.sd = rw_sd(
+    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
+  ),
+  params = params
+) -> mif_out
+
 ```
@@
 ```{r}
+statenames <- c('logF', 'logB')
+paramnames <- c('a', 'b', 'c', 'gamma', 'alpha', 'Beta', 'sigmaF', 'sigmaB',
+                'logF_0', 'logB_0', 'logRho', 'sigma_obs')
+obs_names <- c('logCPUE')
+# logF_0 = 0 normalizing to assume unit initial population
+# sigmaF roughly the noise in a year
+param_guess <- c(
+  logF_0 = 0,
+  logB_0 = 0,
+  a = 1,
+  b = 2,
+  c = 1,
+  alpha = 1,
+  Beta = 1,
+  gamma = 0.5,
+  sigmaF = 0.1,
+  sigmaB = 0.1,
+  logRho = 3,
+  sigma_obs = 0.1
+)
+Csnippet('
+  double dwB, dwF;
+  dwF = rgammawn(sigmaF, dt);
+  dwB = rgammawn(sigmaB, dt);
+  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
+  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
+') -> rproc
+Csnippet('
+  logF = logF_0;
+  logB = logB_0;
+') -> rinit
+Csnippet('
+  log.CPUE = rnorm(logB + logrho, signma_obs);
+') -> rmeas
+Csnippet('
+  lik = dnorm(logCPUE, logB + logrho, sigma_obs, give_log);
+') -> dmeas
+
+my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')
+
+pomp(data = data_frame |> dplyr::select(-R),rprocess=euler(rproc,delta.t=1/12),
+     times = 'year', t0 = 1871,
+    rinit=rinit,
+    # rmeasure=rmeas,
+    # dmeasure=dmeas,
+    statenames=statenames,
+    paramnames=paramnames,
+    covar = my_covars
+  ) -> measSIR
 
 ```
-
-# Methods
-
-# Results
-
-# Conclusion

### 2024-04-23 draft_heyuan.Rmd
Add files via upload

--- 2024-04-23 draft.Rmd
@@
 library(cowplot)
+library(pomp)
+library(foreach)
+library(doFuture)
+plan(multisession)
 library(doParallel)
-library(pomp)
+library(doRNG)
 ```
@@
 ```{r data_table, echo=FALSE}
-df <- read_excel("./data/bird_data.xlsx")
+df <- read_excel("/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/bird_data.xlsx")
 df |>
@@
 
-Looking at both *CPUE* and *log.CPUE* in [figure], it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
+Looking at both *CPUE* and *log.CPUE* plot above, it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
 The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
 It's also worth noting that this trend is far less apparent beyond the early 1900s.
-
-Figure [] depicts the ACF plot of *log.CPUE* at 100 lags.
-The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
-More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
-The PACF plot in figure [], the partial correlation dies out after 5 lags.
-For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
-
-Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data.
-Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods.
-The repeated short-term cycle in 5 lags could be explained by the partial correlation.
 
@@
 ```
+
 ## {-}
+
+Above figure depicts the ACF plot of *log.CPUE* at 100 lags.
+The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
+More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
+The PACF plot above, the partial correlation dies out after 5 lags.
+For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
+
+Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data.
+Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods, which indicates the non-existence of seasonal pattern.
+The repeated short-term cycle in 5 lags could be explained by the partial correlation.
+
 --------
@@
 $$AIC=-2\times\ell(\theta^*)+2D$$
+
+ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for stationarity check, though ACF plot shows an abvious non-stationarity.  
+```{r stationary, echo=FALSE}
+# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
+kpss_result <- kpss.test(ts_cpue, null = "Level")
+print(kpss_result)
+
+# Differencing the time series once
+start_year <- min(data$year)
+end_year <- max(data$year)
+frequency <- 1  
+ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
+ts_diff <- diff(ts_cpue, differences = 1)
+print(kpss.test(ts_diff,null = "Level"))
+```
+
+KPSS with p-value = 0.01 suggests that the average daily hunting of birds(CPUE) in log scale is non-stationary.
+However, first-order difference on this time series data is stationary, proved by the p-value larger than $\alpha=0.05$ from KPSS test.
+Then the following ARMA(p,q) model will be run on such differenced data.
 
@@
 ```
+
+The smaller AIC, the better the model to fit the data.
+Thus, ARMA(1,5) with the lowest AIC value of 217.08 is the best model to fit the differenced data.
+Or in other words, ARIMA(1,1,5) is the best model to fit average daily hunting of birds(CPUE) in log scale.
 
@@
 
-```{r, echo=FALSE, eval=FALSE}
-df |>
-  rename(R = "peak rodent year") |>
-  rename(NAO = "NAO May_Jun_Jul") |>
-  rename(logCPUE = "log.CPUE") |>
-  mutate(R = ifelse(R == "no", 0, 1))  |>
-  rename(reports = "CPUE") |>
-  select(c("year", "reports", "R")) -> bird
-```
-
-```{r echo=FALSE, eval=FALSE}
-birdMod |>
-  pfilter(Np=5, params = params) -> pf
-
-mif2(
-  measSIR, 
-  Nmif = 10, 
-  Np = 2000,
-  cooling.fraction.50 = 0.5, 
-  rw.sd = rw_sd(
-    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
-  ),
-  params = params
-) -> mif_out
-```
-
 ### Local Search
@@
 
-```{r}
+```{r, echo=FALSE, eval=FALSE}
+data_path <- '/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/data.txt'
+data <- read.table(data_path, header = TRUE, sep = "\t", na.strings = "NA", dec = ".", strip.white = TRUE)
+
+data |>
+  rename(R = "peak.rodent.year") |>
+  rename(NAO = "NAO.May_Jun_Jul") |>
+  rename(logCPUE = "log.CPUE") |>
+  mutate(R = ifelse(R == "no", 0, 1)) -> data
+
+subset <- c("year", "logCPUE", "R")
+
 bird <- data[subset]
@@
 
+Fisrt, we used a particle filter applied to a model over time and then assessed the performance.
+We gave a guess of parameter values with the following: $log(F_0) = 1, log(B_0) = 1, a = 1, b = 2, c = 1, \alpha = 1, \beta = 1, \gamma = 0.5, \sigma_F = 0.1,  \sigma_B = 0.1, log(\rho) = 3, \sigma_{obs} = 0.1$
+
 ```{r}
@@
 )
+
+mod |>
+  pfilter(Np=50, params =param_guess) -> pf
+plot(pf)
 ```
 
+The effective sample size (ESS) in the second panel displays a presence of occasional spikes interspersed with predominantly low values.
+It suggests that only a few samples are providing meaningful information, which lead to a less robust and a less reliable estimate.
+The third panel shows the conditional log-likelihood over time which is a measure of how well the model with its current parameters.
+The log-likelihood appears to be fluctuating and showing some periods of higher values interspersed with periods of significant decline, indicating varying model performance over time.
+
+Thus, our initial guess of parameter values is poor and we will do a local search of parameters for a more stablized model performance.
+
+A single execution of the mif2 function applied to mod provides a snapshot of the model's behavior under a specific set of initial conditions.
 ```{r}
+## local search using if2
+## a single execution of the mif2 function applied to mod
 mif2(
@@
 ) -> mif2_out
+
+plot(mif2_out)
+```
+
+We used a foreach loop which iterates 20 times, each iteration of the loop applies the mif2 function to our model.
+This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as control the robustness.
+
+```{r}
+foreach(i=1:20,.combine=c,
+    .options.future=list(seed=482947940)
+  ) %dofuture% {
+    mod |>
+      mif2(
+        Np=1000, Nmif=50,
+        cooling.fraction.50=0.5,
+        rw.sd = rw_sd(
+           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
+           a = 0.02, b = 0.02, c = 0.02, 
+          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
+         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
+        ),
+        partrans = parameter_trans(
+             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
+              'c', 'logF_0', 'logB_0', 'logRho')),
+        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
+                "logF_0", "logB_0", "logRho", "sigma_obs"),
+        params = param_guess
+      )
+  } -> mifs_local
+
+mifs_local |>
+  traces() |>
+  melt() |>
+  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
+  geom_line()+
+  guides(color="none")+
+  facet_wrap(~name,scales="free_y")
 ```
@@
 ```{r}
-freeze(
-  runif_design(
-    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
-              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
-              sigmaF = 0, sigmaB = 0.02, logRho = 0)
-    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
-              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
-              sigmaF = 0, sigmaB = 0.02, logRho = 0)
-  ),
-  seed = 420
-) -> guesses
+foreach(mf=mifs_local,.combine=rbind,
+  .options.future=list(seed=900242057)
+) %dofuture% {
+   evals <- replicate(10, logLik(pfilter(mf,Np=500))) 
+   ll <- logmeanexp(evals,se=TRUE) 
+   mf |> coef() |> bind_rows() |>
+   bind_cols(loglik=ll[1],loglik.se=ll[2]) } -> results
+
+pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results,pch=16)
 ```
@@
 
---------
+```{r}
+
+guesses <- freeze(
+  runif_design(
+    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
+              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
+              sigmaF = 0, sigmaB = 0.02, logRho = 0),
+    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
+              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
+              sigmaF = 0, sigmaB = 0.02, logRho = 0),
+  nseq=400 )
+)
+
+```
+
+```{r}
+set.seed(2062379496)
+
+runif_design(
+  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
+              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
+              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
+    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
+              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
+              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
+    nseq=400
+) -> guesses
+
+mf1 <- mifs_local[[1]]
+fixed_params <- coef(mif2_out)
+
+foreach(guess=iter(guesses,"row"), .combine=rbind,
+      .options.future=list(seed=1270401374)
+    ) %dofuture% {
+      mf1 |>
+        mif2(params=c(guess,fixed_params)) |>
+        mif2(Nmif=1000) -> mf
+      replicate(
+        10,
+        mf |> pfilter(Np=500) |> logLik()
+      ) |>
+        logmeanexp(se=TRUE) -> ll
+      mf |> coef() |> bind_rows() |>
+        bind_cols(loglik=ll[1],loglik.se=ll[2])
+    } -> results
+
+filter(is.finite(loglik)) -> results
+
+```
 

### 2024-04-23 draft.Rmd
small changes

@@
 ```{r}
+
+```
+
+```{r}
 freeze(
@@
               sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
-              sigmaF = 0, sigmaB = 0.02, logRho = 0)
+              sigmaF = 0, sigmaB = 0.02, logRho = 0),
     upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,

### 2024-04-23 draft.Rmd
again

@@
 
-```{r}
-freeze(
-  runif_design(
-    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
-              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
-              sigmaF = 0, sigmaB = 0.02, logRho = 0),
-    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
-              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
-              sigmaF = 0, sigmaB = 0.02, logRho = 0)
-  ),
-  seed = 420
-) -> guesses
-```
-
 ### Global Search

### 2024-04-24 draft.Rmd
big edits

@@
 
-This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan *Lapagos lapagos* in Northeastern Scandinavia ().
+This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan *Lapagos lapagos* in Northeastern Scandinavia [@olav].
 Although not in immediate danger, the willow ptarmigan faces threats from habitat loss, climate change, and hunting and trapping.
-Hjeljord and Loe postulate that, along with climate, a long-term dampening of the amplitude in small rodents and an increase in red fox numbers, have prevented willow ptarmigan populations from reaching their peaks seen a hundred years ago ().
+Hjeljord and Loe postulate that, along with climate, a long-term dampening of the amplitude in small rodents and an increase in red fox numbers, have prevented willow ptarmigan populations from reaching their peaks seen a hundred years ago [@olav].
 Their analysis implements linear models with a count proxy as a function of time to estimate linear change and wavelet analysis to detect cyclic periods.
-This report’s aim is to further Hjelford and Loe’s work by capturing the stochastic population dynamics and the role of alternative prey using the partially observed markov process (POMP) framework
+This report’s aim is to further Hjelford and Loe’s work by capturing the stochastic population dynamics and the role of alternative prey using the partially observed markov process (POMP) framework.
+POMP models allow researchers to make inferences about the underlying dynamics of the system by linking observed and unobservable variables.
+Applications of these models have been carried out extensively in finance and epidemiology, but far less so in ecological systems.
 
@@
 
-Surprisingly, grouse, the family of birds which willow ptarmigan belong to, are of the most studied bird species in the world.
+Perhaps a surprise to some, grouse, the family of birds which willow ptarmigan belong to, are of the most studied bird species in the world.
 Thus, an abundance of scientific research has been carried out to shed light on the decline in its population.
@@
 Studies by Hebbelstrup @hebbelstrup and Storass @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.
-Breisjøberget el al., found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis.
-Despite there being a plethora of work conducted in this area, none have applied
+Breisjøberget el al.
+@breis, found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis.
+Despite there being a plethora of work conducted in this area, none (to our knowledge) have applied a hidden Markov process in attempt to model the population dynamics.
 
 The remainder of the report is structured as follows.
-We briefly introduce and describe our data set and variables of interest.
-Followed by a short exploration of trend, correlation, and association.
-Next, we build a baseline ARMA process model to serve as a measurement for our more complex proposals.
+We briefly introduce and describe our data set and variables of interest followed by a short exploration of trend and correlation.
+Next, we implement a baseline ARMA process model to serve as a measurement for our more complex proposals.
 We then introduce the pomp framework and present our model along with a section on model diagnostics.
-We conclude with the results, a discussion, and a summary with suggestions of future extensions.
+And conclude with the results, a discussion, and a summary with suggestions of future extensions.
 
@@
 
-Our data includes 142 observations of CPUE from 1872 to 2012.
-The harvest data was recorded by hunters in the Southeastern mountain regions of Norway.
+Our data includes 142 observations from 1872 to 2012.
+The harvest data (CPUE) was recorded by hunters in the Southeastern mountain regions of Norway.
 The data is provided by the authors of the study and is hosted by Dryad [here]().
@@
 library(cowplot)
+library(pomp)
 library(doParallel)
-library(pomp)
+library(doRNG)
+library(foreach)
+library(doFuture)
 ```
@@
 
-```{r, echo=FALSE, message=FALSE}
+```{r trend, fig.cap="Full Data Set.", echo=FALSE, message=FALSE}
 # Most basic bubble plot
@@
 
-```{r, boxplot, echo=FALSE}
+```{r boxplot, fig.cap="Boxplot of logCPUE vs Peak Rodent Year.", echo=FALSE}
 ggplot(data, aes(x = factor(R), y = logCPUE, group = factor(R))) + 
@@
 
-Looking at both *CPUE* and *log.CPUE* in [figure], it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
-The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
+Looking at both *CPUE* and *log.CPUE* in Figure \@ref(fig:trend), it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
+The blue shaded regions in plot **C** highlight years when peak rodent year is equal to 1 signifying *yes*.
+Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
 It's also worth noting that this trend is far less apparent beyond the early 1900s.
+Figure \@ref(fig:boxplot) displays side-by-side boxplots of *logCPUE* vs.
+*peak rodent year*, confirming that ptarmigan counts are generally higher in years of rodent abundance.
+**Note:** in accordance with the previous study, our analysis will use *log.CPUE* to reduce heteroscedasticity.
 
-Figure [] depicts the ACF plot of *log.CPUE* at 100 lags.
+Figure \@ref(fig:acf) depicts the ACF plot of *log.CPUE* at 100 lags.
 The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
 More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
-The PACF plot in figure [], the partial correlation dies out after 5 lags.
+The PACF plot in Figure \@ref(fig:trend), the partial correlation dies out after 5 lags.
 For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
@@
 The repeated short-term cycle in 5 lags could be explained by the partial correlation.
-
-In accordance with the previous study, our analysis will use *log.CPUE* to reduce heteroscedasticity.
+That being said, the purpose of this report is to explore the pomp framework, thus we elect to use a very simple ARMA process to serve as a basis for our more sophisticated models.
 
@@
 ### ACF
-```{r ACF, echo=FALSE}
+```{r acf, fig.cap="ACF of logCPUE.", echo=FALSE}
 acf(data$logCPUE, lag = 100, main = "ACF log.CPUE")
@@
 
-```{r PACF, echo=FALSE}
+```{r pacf, fig.cap="ACF of logCPUE.", echo=FALSE}
 pacf(data$logCPUE, lag = 100, main = "PACF log.CPUE")
 ```
+
 ## {-}
+
 --------
@@
 In theory, the ARMA model will serve as a yardstick for managing expectations.
-Of the models shown in (table #), we select the one with the Akaike’s information criterion (AIC).
-AIC is essentially “minus twice the maximized log likelihood plus twice the number of parameters [cite],” and is defined by:
+Of the models shown in Table \@ref(tab:grid), we select the one with the Akaike’s information criterion (AIC).
+AIC is essentially minus twice the maximized log likelihood plus twice the number of parameters, and is defined by:
 
@@
 
-```{r grid_search, echo=FALSE, warning=FALSE}
+We choose the ARMA(3,2), as out simple benchmark model. The corresponding log-likelihood value comes out to be **-105.93**.
+
+```{r grid,echo=FALSE, warning=FALSE}
 aic_table <- function(data,P,Q){
@@
 require(knitr)
-kable(bird_aic_table,digits=2)
+kable(bird_aic_table,digits=2,caption = "AIC Table.")
+```
+
+```{r, eval=FALSE, echo=FALSE}
+library(forecast)
+
+bird_ts <- ts(bird$logCPUE)
+arma_model <- Arima(bird_ts, order = c(3, 0, 2))
+
+# Compute log-likelihood
+log_likelihood <- logLik(arma_model)
+# Check if the model has been fully estimated
+if (!is.na(logLik(arma_model))) {
+  # Compute log-likelihood
+  log_likelihood <- logLik(arma_model)
+  print(log_likelihood)
+} else {
+  print("Model estimation failed. Check model summary for errors.")
+  print(summary(arma_model))
+}
 ```
@@
 
-Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models "consist of an unobserved Markov state process, connected to the data via an explicit model of the observation process." The former is referred to as the *latent process model* and the latter as the *measurement model.* A simple visual representation of the process is depicted in figure [].
+Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable.
+It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data.
+By incorporating both observed and unobserved variables, POMP models allow researchers to make inferences about the underlying dynamics of the system.
+A simple visual representation of the process is depicted in the image below.
 
-We propose modeling the pomp framework using a variation of the Lotka–Volterra equations [cite].
+We propose modeling the pomp framework using a variation of the Lotka–Volterra equations [@lotka].
 The Lotka-Voltera equations, also referred to as the predator-prey equations, are a pair of first order nonlinear differential equations.
@@
 Our analysis focuses on a slightly different approach that introduces the idea of an *alternative prey.* The alternative prey hypothesis supposes that predators supported by a primary species will shift to consume alternative prey during a decrease in primary prey abundance.
-
-The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator (fox), preferred prey(rodent) and alternative prey (bird) count.
+Despite this occurring in many systems, the mechanisms are poorly understood [@brunet], and so we offer an avenue to explore by introducing the pomp framework.
 
@@
 
-### Model Components
+The measurement model $Y(t)$ is our ptarmigan count proxy, *logCPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator (fox) population density, preferred (rodent) prey population density, and alternative prey (bird) population density.
+We model how the log of the populations of foxes and ptarmigans change through time in Equations \@ref(eq:fox) and \@ref(eq:bird).
 
-- `rprocess`: a draw from $f_{X_n|X_{n-1}}(x_n|x_{n-1};\theta)$
-- `dprocess`: evaluation of $f_{X_n|X_{n-1}}(x_n|x_{n-1};\theta)$
-- `rmeasure`: a draw from $f_{Y_n|X_{n}}(y_n|x_{n};\theta)$
-- `dmeasure`: evaluation of $f_{Y_n|X_{n}}(y_n|x_{n};\theta)$
-- `rinit`:a draw from $f_{X_0}(x_0;\theta)$
+\begin{equation}
+\log F_{t+dt} = \log F_t + dt(bR_t + c\exp(\log B_t)[1 - \gamma R_t] - a)W_t^F (\#eq:fox)
+\end{equation}
 
-**Covariates**
+\begin{equation}
+\log B_{t+dt} = \log B_t + dt(\alpha + \beta \exp(\log F_t)[1 - \gamma R_t])W_t^F (\#eq:bird)
+\end{equation}
 
-- $Z(t)= R(t)$ peak rodent yea
+In equation \@ref(eq:fox):
 
-$$Y_n = N$$
+- $\log F_{t+dt}$ is the change in the log of the fox population density over a small time interval $dt$;
+- $\log F_t$ is the current log fox population density;
+- $bR_t$ represents the fox reproduction rate influenced by the rodent population density $R(t)$;
+- $a$ represents the fox death rate;
+- $c\exp(\log B_t)[1−\gamma Rt]$ represents the predation rate of foxes on birds, where $c$ is a parameter representing the efficiency of fox predation, $\exp(\log B_t)$ is the bird population density, $\gamma$ is a parameter representing the impact of rodent population on predation efficiency, and $[1−\gamma R_t]$ accounts for the reduction in predation efficiency when the rodent population is high.
+- $W_t^F$ is an integrated gamma white noise variable;
+- $t$ is time.
 
-$$Y_t \sim NB(\rho B_t, \sigma),$$
-where parameter $\rho$ is proportional to capturing effort, e.g., the number of hunters.
+Similarly, in equation \@ref(eq:bird):
 
-$$F_{t+dt}=\delta(t)F_tB_t-\theta F_t$$ 
-where
-
-$$
-\delta (t) = \left\{\begin{array}{ll}
-\delta_1 & : R(t)=0\\
-\psi \delta_1 & :  o.w.
-\end{array}
-\right.
-$$
-
-$\delta(t)F_t\beta_t$ is the rate of the poisson step.
-
-$$B_{t+dt} = \alpha B_t - \beta(R_t)B_tF_t,$$
-
-where
-
-$$
-\beta (R_t) = \left\{\begin{array}{ll}
-\beta_1 & : R(t)=0\\
-\gamma \beta_1 & :  o.w.
-\end{array}
-\right.
-$$
-
-We use a binomial approximation with transition probabilities
+- $\log B_{t+dt}$ is the change in the log of the ptarmigan population density over a small time interval $dt$;
+- $\log B_t$ is the current log ptarmigan population density;
+- $\alpha$ and $\beta$ represent ptarmigan birth rate and the effect of fox predation on ptarmigan population, respectively;
+- $\exp \log(F_t)$ represents the fox population density influencing bird predation;
+- $W_t^F$ and $t$ represent the same interpretations as in Equation \@ref(eq:fox).
 
@@
 
-```{r echo=FALSE, eval=FALSE}
-birdMod |>
-  pfilter(Np=5, params = params) -> pf
-
-mif2(
-  measSIR, 
-  Nmif = 10, 
-  Np = 2000,
-  cooling.fraction.50 = 0.5, 
-  rw.sd = rw_sd(
-    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
-  ),
-  params = params
-) -> mif_out
-```
-
 ### Local Search
 
-```{r}
+We start by fitting our model to the data using a local search.
+Code for the model can be viewed using the black drop-down boxes throughout the report.
+Version 5.6 of the `pomp` package was used to carry out the analysis.
+We use a particle filter applied to a pomp object over time and assess its performance.
+We initialize the process with a guess of parameter values given by the following: $log(F_0) = 1, log(B_0) = 1, a = 1, b = 2, c = 1, \alpha = 1, \beta = 1, \gamma = 0.5, \sigma_F = 0.1,  \sigma_B = 0.1, log(\rho) = 3, \sigma_{obs} = 0.1$
+
+```{r, echo=FALSE}
 statenames <- c("logF", "logB")
@@
 
-```{r}
+```{r, echo=FALSE}
 bird <- data[subset]
-head(bird)
 ```
@@
 
-```{r}
+```{r, echo=FALSE}
 bird1 <- bird |> 
@@
     )
-    # covarnames <- c("R")
   ) -> mod
@@
 
-```{r}
+```{r, echo=FALSE}
 param_guess <- c(
   logF_0 = 1, 
-  logB_0 = 1,
+  logB_0 = 2,
   a = 1,
@@
 
-```{r}
+```{r pf1, fig.cap="Particle Pilter with Np=50", echo=FALSE}
+mod |>
+  pfilter(Np=50, params =param_guess) -> pf
+plot(pf)
+```
+
+The effective sample size (ESS) in the second panel of Figure \@ref(fig:pf1) displays a presence of occasional spikes interspersed with predominantly low values.
+It suggests a particle degeneracy issue.
+In other words, only a few samples are providing meaningful information, which lead to a less robust and a less reliable estimate.
+The third panel shows the conditional log-likelihood over time which is a measure of how well the model with its current parameters.
+The log-likelihood appears to be fluctuating and showing some periods of higher values interspersed with periods of significant decline, indicating varying model performance over time.
+We can conclude that our initial guess of parameter values is poor and we will do a local search of parameters for a more stable model performance.
+
+A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions.
+
+```{r, echo=FALSE, eval=FALSE}
 mif2(
@@
 
-```{r}
+Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model.
+This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as to control the robustness.
+The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.
 
+```{r, echo=FALSE, eval=FALSE}
+bake(file="local_search.rds", {
+  foreach(i=1,.combine=c,
+    .options.future=list(seed=482947940)
+  ) %dofuture% {
+    mod |>
+      mif2(
+        Np=1000, Nmif=50,
+        cooling.fraction.50=0.5,
+        rw.sd = rw_sd(
+           alpha = 0.02, Beta = 0.02, gamma = 0.02, 
+           a = 0.02, b = 0.02, c = 0.02, 
+          sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
+         sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
+        ),
+        partrans = parameter_trans(
+             log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
+              'c', 'logF_0', 'logB_0', 'logRho')),
+        paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
+                "logF_0", "logB_0", "logRho", "sigma_obs"),
+        params = param_guess
+      )
+  }}) -> mifs_local
+
+mifs_local |>
+  traces() |>
+  melt() |>
+  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
+  geom_line()+
+  guides(color="none")+
+  facet_wrap(~name,scales="free_y")
+
+bake(file="lik_local.rds",{
+  foreach(mf=mifs_local,.combine=rbind,
+    .options.future=list(seed=900242057)
+  ) %dofuture% {
+    evals <- replicate(10, logLik(pfilter(mf,Np=100)))
+    ll <- logmeanexp(evals,se=TRUE)
+    mf |> coef() |> bind_rows() |>
+      bind_cols(loglik=ll[1],loglik.se=ll[2])
+  } -> results
+  attr(results,"ncpu") <- nbrOfWorkers()
+  results
+}) -> results_local
+
+bind_rows(results_local) |> arrange(-loglik) |> write_csv("bird_params.csv")
+
+if (file.exists("CLUSTER.R")) {
+  source("CLUSTER.R")
+}
 ```
 
+The likelihood over iterations quickly rises and then levels off.
+Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out.
+This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty.
+But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.
+
+We find that the higher likelihood is usually associated with lower values of $\alpha, \gamma, log(\rho), \sigma_{obs}$.
+$log(F_0), log(B_0)$, the initial conditions for Fox population and birds population in log scale, shows a general upward trend over iterations.
+$a, b, c, \sigma_F, \sigma_B$ shows a general flat trend near zero except for a few iterations with a dramatically increasing trend.
+
 ### Global Search
-
---------
-
-## Diagnostics
 
@@
 
+Our report probes the potential of applying POMP to the field of ecological population dynamics.
+We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models.
+Drawing inspiration from the Lotka-Voltera equations, we introduce our take on the concept of an alternative prey hypothesis.
+Although our analysis is on the side of preliminary, it shows this approach can shed more light on the dynamic responses of ecosystems to changes in prey abundance.
+
 --------
@@
 - Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
-- Heyuan: ?
+- Heyuan: 
 - Sizhuang:

### 2024-04-24 draft_heyuan.Rmd

@@
 
-Looking at both *CPUE* and *log.CPUE* plot above, it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
+Looking at both *CPUE* and *log.CPUE* plot above with decreasing number of hunting yield, it's evident that ptarmigan populations has steadily decreased over the 142 year-long period.
 The blue shaded regions in plot **C** highlight years when peak rodent year is equal to "yes." Notice how both plots exhibit peaks in the shaded regions and troughs otherwise, which is exactly what we expected to see from the alternative prey hypothesis.
@@
 
+$$F_{t+dt} = bR + ce^{logB}* (1-\gamma R)-a + F_{t}$$
+
+$$B_{t+dt} = \alpha - \beta e^{logF} *(1-\gamma R) + B_t$$
+
+$logF_0$ and $logB_0$ are fox population and bird population in log scale in the initial time 0. 
+$R$ is a binary variable with 0(no) or 1(yes), indicating whether this year is a peak rodent year or not.
+$F_{dt}(dF)$ can be generated from the gamma distribution with $\sigma_F$ and a timestep $dt$. $B_{dt}(dB)$ can be generated from the gamma distribution with $\sigma_B$ and a timestep $dt$.
+Our observation log CPUE follows the normal distribution with mean $logB - log(\rho)$ and standard deviation $\sigma_{obs}$.
+We will use the stochastic Euler's Approximation method to built our POMP model by Csnippet.
+
 ```{r}
@@
 The effective sample size (ESS) in the second panel displays a presence of occasional spikes interspersed with predominantly low values.
-It suggests that only a few samples are providing meaningful information, which lead to a less robust and a less reliable estimate.
+It suggests a particle degeneracy issue.
+In other words, only a few samples are providing meaningful information, which lead to a less robust and a less reliable estimate.
 The third panel shows the conditional log-likelihood over time which is a measure of how well the model with its current parameters.
@@
 
-We used a foreach loop which iterates 20 times, each iteration of the loop applies the mif2 function to our model.
+Iterated Filtering (IF2) is implemented by a foreach loop which iterates 20 times, and each iteration of the loop applies the IF2 function to our model.
 This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as control the robustness.
+The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.
 
@@
 
-```{r}
-foreach(mf=mifs_local,.combine=rbind,
-  .options.future=list(seed=900242057)
-) %dofuture% {
-   evals <- replicate(10, logLik(pfilter(mf,Np=500))) 
-   ll <- logmeanexp(evals,se=TRUE) 
-   mf |> coef() |> bind_rows() |>
-   bind_cols(loglik=ll[1],loglik.se=ll[2]) } -> results
+The likelihood over iterations quickly rises and then levels off.
+Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out.
+This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty.
+But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.
 
-pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results,pch=16)
-```
+I find that the higher likelihood is usually associated with lower values of $\alpha, \gamma, log(\rho), \sigma_{obs}$.
+$log(F_0), log(B_0)$, the initial conditions for Fox population and birds population in log scale, shows a general upward trend over iterations.
+$a, b, c, \sigma_F, \sigma_B$ shows a general flat trend near zero except for a few iterations with a dramatically increasing trend.
 
@@
 
+```{r, echo=FALSE}
+print(paste('The maximum likelihood from global search for our model is', results$loglik, 
+            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))
+
+best_index <- which.max(results$loglik)
+best_parameters <- results[best_index, ]
+print(paste('The best parameters are:', toString(best_parameters), 'for beta, rho, eta, mu_EI, mu_IR,'))
+```
+
 ## Diagnostics
+
+#### ARMA
+
+```{r}
+fit1 <- arima(ts_cpue, order=c(0,0,5))
+residuals <- residuals(fit1)
+
+# ACF check
+acf(residuals, main="ACF of Residuals")
+
+# Normality check with Q-Q plot
+qqnorm(residuals)
+qqline(residuals, col="red")
+
+# Shapiro-Wilk test
+shapiro.test(residuals)
+
+# Root Check
+ma_coefficients <- coef(fit1)[grepl("ma", names(coef(fit1)))]
+MA_roots <- polyroot(c(1, ma_coefficients))
+#invertible <- all(Mod(MA_roots) > 1)
+#print(paste("Is the model invertible? ", invertible))
+
+autoplot(fit1) +
+  ggtitle("ARMA(0, 5)")
+```
+
+For the ACF plot of the residuals, all autocorrelations are within the confidence bounds (indicated by blue dashed lines), which suggests that there is no significant autocorrelation present.
+Additionally, the QQ plot shows that the data points roughly align with the reference line (colored red), supporting the assumption of normality.
+This is further confirmed by a Shapiro-Wilk test for normality, which returns a p-value of 0.62, indicating that the residuals can be considered normally distributed.
+Moreover, the roots of the characteristic polynomial are outside the unit circle, confirming that the MA(5) process is invertible — a critical condition for model adequacy.
+These diagnostics collectively verify that the ARIMA(1,1,5) model is well-specified and adheres to the necessary assumptions for valid time series analysis.
 

### 2024-04-24 draft.Rmd
local searches

@@
 
-Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable.
-It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data.
+Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable [@ch10].
+It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data [@start].
 By incorporating both observed and unobserved variables, POMP models allow researchers to make inferences about the underlying dynamics of the system.
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 bird1 <- bird |> 
@@
 
-A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions.
-
 ```{r, echo=FALSE, eval=FALSE}
@@
 
+A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions.
 Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model.
@@
 
+```{r, message=FALSE}
+# bake(file="local_search.rds", {
+#   foreach(i=1:20,.combine=c,
+#     .options.future=list(seed=482947940)
+#   ) %dofuture% {
+#     mod |>
+#       mif2(
+#         Np=1000, Nmif=50,
+#         cooling.fraction.50=0.5,
+#         rw.sd = rw_sd(
+#            alpha = 0.02, Beta = 0.02, gamma = 0.02,
+#            a = 0.02, b = 0.02, c = 0.02,
+#           sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
+#          sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
+#         ),
+#         partrans = parameter_trans(
+#              log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
+#               'c', 'logF_0', 'logB_0', 'logRho')),
+#         paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB",
+#                 "logF_0", "logB_0", "logRho", "sigma_obs"),
+#         params = param_guess
+#       )
+#   }}) -> mifs_local
+
+mifs_local |>
+  traces() |>
+  melt() |>
+  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
+  geom_line()+
+  guides(color="none")+
+  facet_wrap(~name,scales="free_y")
+```
+
 The likelihood over iterations quickly rises and then levels off.
@@
 $a, b, c, \sigma_F, \sigma_B$ shows a general flat trend near zero except for a few iterations with a dramatically increasing trend.
+
+```{r}
+# bake(file="lik_local.rds",{
+#   foreach(mf=mifs_local,.combine=rbind,
+#     .options.future=list(seed=900242057)
+#   ) %dofuture% {
+#     evals <- replicate(10, logLik(pfilter(mf,Np=100)))
+#     ll <- logmeanexp(evals,se=TRUE)
+#     mf |> coef() |> bind_rows() |>
+#       bind_cols(loglik=ll[1],loglik.se=ll[2])
+#   } -> results
+#   attr(results,"ncpu") <- nbrOfWorkers()
+#   results
+# }) -> results_local
+# bind_rows(results_local) |> arrange(-loglik) |> write_csv("bird_params.csv")
+# if (file.exists("CLUSTER.R")) {
+#   source("CLUSTER.R")
+# }
+pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
+```
 

### 2024-04-24 blinded.Rmd
ready version

--- 2024-04-24 draft.Rmd
@@
 title: "Investigating the alternative prey hypothesis with the POMP framework"
-author: "Pete, Sizhuang, and Heyuan"
+author: "Blinded"
 date: "`r Sys.Date()`"
@@
 library(pomp)
+plan(multisession)
 library(doParallel)
@@
 
-Figure \@ref(fig:acf) depicts the ACF plot of *log.CPUE* at 100 lags.
-The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
-More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
-The PACF plot in Figure \@ref(fig:trend), the partial correlation dies out after 5 lags.
-For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
-
-Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data.
-Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods.
-The repeated short-term cycle in 5 lags could be explained by the partial correlation.
-That being said, the purpose of this report is to explore the pomp framework, thus we elect to use a very simple ARMA process to serve as a basis for our more sophisticated models.
-
 ## Correlation {.tabset}
@@
 
+Figure \@ref(fig:acf) depicts the ACF plot of *log.CPUE* at 100 lags.
+The plot reinforces what we already know - the time series is not stationary, or that future values of the time series are correlated with past values.
+More specifically, the initial 40 lags show strong correlation with the strongest being the first 5.
+The PACF plot in Figure \@ref(fig:trend), the partial correlation dies out after 5 lags.
+For an ARMA model, it may be appropriate to select an $AR(p=5)$ component.
+
+Given the long-range significance in the ACF plot and obvious non-stationarity, differencing the series may be necessary to fit a simple time series model to the data.
+Moreover, the ACF appears to be approaching a seasonal behavior, however we cannot apply STL or classical decomposition procedures as our sample of data has less than 2 periods.
+The repeated short-term cycle in 5 lags could be explained by the partial correlation.
+That being said, the purpose of this report is to explore the pomp framework, thus we elect to use a very simple ARMA process to serve as a basis for our more sophisticated models.
+
 --------
@@
 
-We choose the ARMA(3,2), as out simple benchmark model. The corresponding log-likelihood value comes out to be **-105.93**.
+ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for a stationarity check.
+```{r stationary, echo=FALSE}
+# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
+kpss_result <- tseries::kpss.test(ts_cpue, null = "Level")
+print(kpss_result)
+
+# Differencing the time series once
+start_year <- min(data$year)
+end_year <- max(data$year)
+frequency <- 1  
+ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
+ts_diff <- diff(ts_cpue, differences = 1)
+print(tseries::kpss.test(ts_diff,null = "Level"))
+```
+
+KPSS with p-value = 0.01 suggests that the average daily hunting of birds(CPUE) in log scale is non-stationary.
+However, first-order difference on this time series data is stationary, proved by the p-value larger than $\alpha=0.05$ from KPSS test.
+Then the following ARMA(p,q) model will be run on such differenced data.
 
@@
 
-bird_aic_table <- aic_table(data$logCPUE,4,5)
+bird_aic_table <- aic_table(ts_diff,4,5)
 require(knitr)
@@
 ```
+
+The best ARMA(0,5) is selected given the lowest AIC with 204.48 to fit first-order differenced data. Thus, ARIMA(0,1,5) will be run below to check for its log-likelihood.
 
@@
 bird_ts <- ts(bird$logCPUE)
-arma_model <- Arima(bird_ts, order = c(3, 0, 2))
+arma_model <- Arima(bird_ts, order = c(0, 1, 5))
 
@@
 
+We choose the ARIMA(0,1,5), as out simple benchmark model. The corresponding log-likelihood value comes out to be **-99.32**.
+
 ## Partially Observed Markov Process
 
-Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable.
-It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data.
+Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable [@ch10].
+It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data [@start].
 By incorporating both observed and unobserved variables, POMP models allow researchers to make inferences about the underlying dynamics of the system.
@@
 ) -> mif2_out
+
+foreach (
+  i=1:10,
+  .combine=c,
+  .options.future=list(seed=652643293)
+) %dofuture% {
+  mif2_out |> pfilter(Np=5)
+} -> pf
+logLik(pf) -> ll
+print(logmeanexp(ll,se=TRUE))
 ```
+
+The estimated log-likelihood is -205 with standard error of 3.14.
 
@@
 bake(file="local_search.rds", {
-  foreach(i=1,.combine=c,
+  foreach(i=1:20,.combine=c,
     .options.future=list(seed=482947940)
@@
   }}) -> mifs_local
+
+logLik(mifs_local) -> ll
+print(logmeanexp(ll,se=TRUE))
 
@@
 
-The likelihood over iterations quickly rises and then levels off.
+It estimated log likelihood of -134 which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
+![Model parameters over iterations](./assets/local_search_1.png)
+From the above plot, the likelihood over iterations quickly rises and then levels off.
 Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out.
@@
 
+```{r}
+pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
+```
+
+![](./assets/local_search_2.png)
+A scattered or cloud-like pattern indicates little to no apparent linear correlation.
+
 ### Global Search
 
---------
+```{r}
+set.seed(2062379496)
 
-# Results
+runif_design(
+  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
+              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
+              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
+    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
+              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
+              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
+    nseq=50
+) -> guesses
 
---------
+mf1 <- mifs_local[[1]]
+fixed_params <- coef(mif2_out)
+
+bake(file="global_search_middle2.rds",
+  dependson=guesses,{
+    foreach(guess=iter(guesses,"row"), .combine=rbind,
+      .options.future=list(seed=1270401374)
+    ) %dofuture% {
+      mf1 |>
+        mif2(params=c(guess,fixed_params)) |>
+        mif2(Nmif=100) -> mf
+      replicate(
+        10,
+        mf |> pfilter(Np=500) |> logLik()
+      ) |>
+        logmeanexp(se=TRUE) -> ll
+      mf |> coef() |> bind_rows() |>
+        bind_cols(loglik=ll[1],loglik.se=ll[2])
+    } -> results
+    attr(results,"ncpu") <- nbrOfWorkers()
+    results
+  }) |>
+  filter(is.finite(loglik)) -> results
+
+bind_rows(results) |>
+  filter(is.finite(loglik)) |>
+  arrange(-loglik) |>
+  write_csv("bird_params_middle2.csv")
+
+```
+
+```{r, echo=FALSE}
+results <- read.csv('/Users/ruojunliu/Desktop/bird_params_small.csv')
+print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
+            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))
+
+best_index <- which.max(results$loglik)
+best_parameters <- results[best_index, ]
+print(paste('The best parameters are:', toString(best_parameters)))
+```
+
+The best estimated log likelihood under global search by the use of GreatLakes is -320.4, which is a lot lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134).
+It is opposite to our expectation.
+One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters.
+The accuracy will be increased if we increase the parameter space.
+
+As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric
 
@@
 Our report probes the potential of applying POMP to the field of ecological population dynamics.
+It ends with ARIMA(0,1,5) produces the highest log-likelihood with **-99**.
 We began with the establishment of a foundational ARMA model to serve as a reference point against which we evaluated our proposed POMP models.
@@
 # Contribution
-
-- Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
-- Heyuan: 
-- Sizhuang:
-
+The contribution is not included in order to be blinded.
 --------

### 2024-04-24 draft.Rmd
ready version

--- 2024-04-24 draft.Rmd
@@
 
+```{r}
+pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
+```
+
+![](./assets/local_search_2.png)
+A scattered or cloud-like pattern indicates little to no apparent linear correlation.
+
 ### Global Search
 
---------
+```{r}
+set.seed(2062379496)
 
-# Results
+runif_design(
+  lower = c(alpha = 0, Beta = 0, gamma = 0, a = 7, b = 2, c = 0.5,
+              sigma_obs = 0, logF_0 = 0.4, logB_0 = 1.6,
+              sigmaF = 0, sigmaB = 0.02, logRho = 0.3),
+    upper = c(alpha = 0.1, Beta = 0.1, gamma = 0.1, a = 8, b = 3, c = 1,
+              sigma_obs = 1, logF_0 = 0.5, logB_0 = 1.7,
+              sigmaF = 0.1, sigmaB = 0.04, logRho = 0.5),
+    nseq=50
+) -> guesses
 
---------
+mf1 <- mifs_local[[1]]
+fixed_params <- coef(mif2_out)
+
+bake(file="global_search_middle2.rds",
+  dependson=guesses,{
+    foreach(guess=iter(guesses,"row"), .combine=rbind,
+      .options.future=list(seed=1270401374)
+    ) %dofuture% {
+      mf1 |>
+        mif2(params=c(guess,fixed_params)) |>
+        mif2(Nmif=100) -> mf
+      replicate(
+        10,
+        mf |> pfilter(Np=500) |> logLik()
+      ) |>
+        logmeanexp(se=TRUE) -> ll
+      mf |> coef() |> bind_rows() |>
+        bind_cols(loglik=ll[1],loglik.se=ll[2])
+    } -> results
+    attr(results,"ncpu") <- nbrOfWorkers()
+    results
+  }) |>
+  filter(is.finite(loglik)) -> results
+
+bind_rows(results) |>
+  filter(is.finite(loglik)) |>
+  arrange(-loglik) |>
+  write_csv("bird_params_middle2.csv")
+
+```
+
+```{r, echo=FALSE}
+results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
+print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
+            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))
+
+best_index <- which.max(results$loglik)
+best_parameters <- results[best_index, ]
+print(paste('The best parameters are:', toString(best_parameters)))
+```
+
+The best estimated log likelihood under global search by the use of GreatLakes is -176.3, which is a lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134).
+It is opposite to our expectation.
+One of the reasons is that we limit the parameter space to save the running time on Great Lakes, otherwise it will take more than 8 hours to run given the fact that our model is built with a large number of parameters.
+The accuracy will be increased if we increase the parameter space.
+
+As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric
 
@@
 
-- Pete: Proposed the research topic and found data set. Wrote introduction, related, work, exploratory data analysis sections. 
-- Heyuan: 
-- Sizhuang:
-
+- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections, ARMA. 
+- Heyuan: Wrote the interpretation of result, refine the ARIMA, do the POMP, local search and global search. 
+- Sizhuang: Helped with POMP, local search and global search and created visualizations of some of the POMP results
+Professor and GSI helped us a lot in building POMP model.
 --------

### 2024-04-24 unblinded.Rmd
ready version

--- 2024-04-24 draft.Rmd
@@
 ```{r, echo=FALSE}
-results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
+results <- read.csv('/Users/ruojunliu/Desktop/bird_params_small.csv')
 print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
@@
 
-The best estimated log likelihood under global search by the use of GreatLakes is -176.3, which is a lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134).
+The best estimated log likelihood under global search by the use of GreatLakes is -320.4, which is a lot lower than the log likelihood from ARIMA (log-likelihood with -99) and local search(log-likelihood with -134).
 It is opposite to our expectation.
@@
 
-- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections, ARMA. 
-- Heyuan: Wrote the interpretation of result, refine the ARIMA, do the POMP, local search and global search. 
+- Pete: Proposed the research topic and found data set. Wrote introduction, related work, exploratory data analysis sections. 
+- Heyuan: Wrote the interpretation of result, do the POMP, local search and global search. 
 - Sizhuang: Helped with POMP, local search and global search and created visualizations of some of the POMP results

### 2024-04-24 workspace/sizhuang.Rmd
ready version

@@
 ```
+
+```{r}
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+# Load data
+data <- read.csv("/Users/sizhuang/Desktop/bird_params_small.csv")
+
+# Reshape data to long format
+data_long <- data %>%
+  pivot_longer(
+    cols = -c(loglik, loglik.se), # Exclude loglik and loglik.se from reshaping
+    names_to = "variable",
+    values_to = "value"
+  )
+
+# Calculate the threshold for the red line
+threshold <- max(data$loglik) - 0.5 * qchisq(p = 0.95, df = 1)
+
+# Define the number of variables per plot
+vars_per_plot <- 5  # Adjust this as needed to fit your specific requirements
+total_vars <- length(unique(data_long$variable))
+total_plots <- ceiling(total_vars / vars_per_plot)
+
+# Create and display/save multiple plots
+plots <- list()
+for (i in 1:total_plots) {
+  subset_vars <- unique(data_long$variable)[((i-1) * vars_per_plot + 1):(i * vars_per_plot)]
+  subset_data <- data_long %>% filter(variable %in% subset_vars)
+
+  # Adjusted for more columns to expand vertical space
+  ncol_facet = min(3, length(subset_vars))  # Use 3 columns or less based on available variables
+  nrow_facet = ceiling(length(subset_vars) / ncol_facet)  # Calculate number of rows based on number of columns
+
+  plots[[i]] <- ggplot(subset_data, aes(x = value, y = loglik)) +
+    geom_point(alpha = 0.3) +
+    geom_hline(yintercept = threshold, color = "red", linetype = "dashed", size = 1) +
+    facet_wrap(~ variable, scales = "free_x", ncol = ncol_facet, nrow = nrow_facet) +
+    labs(title = paste("Variables vs Loglik - Plot", i),
+         x = "Value",
+         y = "Loglik") +
+    theme_minimal() +
+    theme(strip.text.x = element_text(size = 12, face = "bold"),
+          legend.position = "none",
+          axis.title.y = element_text(size = 14),
+          axis.title.x = element_text(size = 14),
+          panel.spacing = unit(1.5, "lines"))
+
+  # Display or save plots
+  print(plots[[i]])
+  # Optionally save each plot
+  # ggsave(paste("Plot_", i, ".png", sep = ""), plot, width = 20, height = 8 * nrow_facet, units = "in")
+}
+
+```

### 2024-04-24 unblinded.Rmd
Add files via upload

--- 2024-04-24 draft.Rmd
@@
 csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
-bibliography: references.bib 
+bibliography: /Users/ruojunliu/Desktop/references.bib 
 output: 
@@
 
-![Female Willow Ptarmigan captured by [Bryce W. Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](./assets/ptarmigan.jpg)
+![Female Willow Ptarmigan captured by [Bryce W.
+Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/ptarmigan.jpg)
 
@@
 library(pomp)
-plan(multisession)
 library(doParallel)
@@
 ```{r stationary, echo=FALSE}
+start_year <- min(data$year)
+end_year <- max(data$year)
+frequency <- 1  
+ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
+
 # Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
@@
 # Differencing the time series once
-start_year <- min(data$year)
-end_year <- max(data$year)
-frequency <- 1  
-ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
 ts_diff <- diff(ts_cpue, differences = 1)
@@
 
-![Conditional independence graph of POMP from A. King's Notes](./assets/state_space.png)
+![Conditional independence graph of POMP from A. King's Notes](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/state_space.png)
 
@@
 
-```{r, echo=FALSE, eval=FALSE}
+```{r, echo=FALSE}
 mif2(
@@
 
-The estimated log-likelihood is -205 with standard error of 3.14.
+The estimated log-likelihood is **-205** with standard error of 3.14.
 
@@
 
+pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
+
 bind_rows(results_local) |> arrange(-loglik) |> write_csv("bird_params.csv")
@@
 
-It estimated log likelihood of -134 which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
-![Model parameters over iterations](./assets/local_search_1.png)
+It estimated log likelihood of **-134** which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
+![Model parameters over iterations](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/local_search_1.png)
+
 From the above plot, the likelihood over iterations quickly rises and then levels off.
@@
 
-```{r}
-pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
+```{r, echo=FALSE, eval=FALSE}
+#pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
 ```
 
-![](./assets/local_search_2.png)
+![](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/local_search_2.png)
+
 A scattered or cloud-like pattern indicates little to no apparent linear correlation.
@@
 
-```{r}
+```{r, echo=FALSE, eval=FALSE}
 set.seed(2062379496)
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, eval=FALSE}
 results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
@@
 
-As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric
+As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric.
 
